\documentclass[aspectratio=169,11pt]{beamer}

\usetheme{Madrid}
\usecolortheme{dolphin}
\setbeamertemplate{navigation symbols}{}
\setbeamertemplate{footline}{%
  \leavevmode%
  \hbox{%
    \begin{beamercolorbox}[wd=.78\paperwidth,ht=2.7ex,dp=1.2ex,leftskip=1.2em]{author in head/foot}%
      \usebeamerfont{author in head/foot}A Journey to the World of Numbers
    \end{beamercolorbox}%
    \begin{beamercolorbox}[wd=.22\paperwidth,ht=2.7ex,dp=1.2ex,rightskip=1.2em plus1fil]{date in head/foot}%
      \usebeamerfont{date in head/foot}\hfill\insertframenumber/\inserttotalframenumber
    \end{beamercolorbox}%
  }%
  \vskip0pt%
}

\definecolor{journeyblue}{RGB}{27,75,125}
\definecolor{journeygold}{RGB}{184,126,23}
\setbeamercolor{structure}{fg=journeyblue}
\setbeamercolor{alerted text}{fg=journeygold}
\setbeamercolor{title}{fg=white,bg=journeyblue}
\setbeamercolor{frametitle}{fg=white,bg=journeyblue}
\setbeamercolor{block title}{fg=white,bg=journeyblue}
\setbeamercolor{block body}{bg=journeyblue!5}
\setbeamercolor{block title alerted}{fg=white,bg=journeygold}
\setbeamercolor{block body alerted}{bg=journeygold!8}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amsmath,amssymb,mathtools}
\usepackage{microtype}

\newcommand{\N}{\mathbb N}
\newcommand{\suc}[1]{S(#1)}
\newcommand{\Z}{\mathbb Z}
\newcommand{\Q}{\mathbb Q}
\newcommand{\R}{\mathbb R}
\newcommand{\C}{\mathbb C}
\newcommand{\cl}[2]{\left[(#1,#2)\right]}
\newcommand{\qcl}[2]{\left[\frac{#1}{#2}\right]}

\title{A Journey to the World of Numbers}
\author{Riccardo Brasca \and Shanwen Wang}
\institute{Renmin University}
\date{July 2026}

\AtBeginSection[]{
  \begin{frame}[plain]
    \centering
    \vfill
    {\usebeamerfont{title}\color{journeyblue}\Huge\insertsectionhead}
    \vfill
  \end{frame}
}

\begin{document}

\begin{frame}[plain]
  \titlepage
\end{frame}

\begin{frame}
  \small
  \tableofcontents[hideallsubsections]
\end{frame}

\section{The natural numbers}

\begin{frame}{Starting from \(0\)}
  The natural numbers form the sequence
  \[
    0,\qquad \suc 0,\qquad \suc{\suc 0},\qquad
    \suc{\suc{\suc 0}},\quad\ldots
  \]
  governed by the following principles.

  \begin{enumerate}
    \item \(0\in\N\).
    \item If \(n\in\N\), then its \alert{successor} \(\suc n\in\N\).
    \item Every natural number is either \(0\) or the successor of a natural number.
    \item No successor equals \(0\): \(\suc n\ne0\).
    \item Successors are unique: \(\suc m=\suc n\) implies \(m=n\).
  \end{enumerate}

  We set \(1=\suc 0\).  Once addition is defined, we will prove that
  \(\suc n=n+1\).
\end{frame}

\begin{frame}{The induction principle}
  \begin{block}{Principle of mathematical induction}
    Let \(P(n)\) be a statement about \(n\in\N\).  If
    \begin{enumerate}
      \item \(P(0)\) is true, and
      \item for every \(n\in\N\), \(P(n)\) implies \(P(\suc n)\),
    \end{enumerate}
    then \(P(n)\) is true for every \(n\in\N\).
  \end{block}

  \vspace{0.4em}
  Induction mirrors the way the natural numbers are generated:
  \[
    \underbrace{P(0)}_{\text{start}}
    \Longrightarrow P(\suc 0)
    \Longrightarrow P(\suc{\suc 0})
    \Longrightarrow \cdots
  \]
\end{frame}

\begin{frame}{The recursion principle}
  \begin{block}{Defining a function on \(\N\)}
    Let \(X\) be a set, choose \(x_0\in X\), and let
    \(G\colon\N\times X\to X\) be a function.  There is one and only one function
    \(f\colon\N\to X\) satisfying
    \[
      f(0)=x_0,
      \qquad
      f(\suc n)=G\bigl(n,f(n)\bigr)
      \quad\text{for every }n\in\N.
    \]
  \end{block}

  \begin{alertblock}{Why this matters}
    Addition, multiplication, and the predecessor function will all be defined
    by this principle.
  \end{alertblock}
\end{frame}

\subsection{Addition}

\begin{frame}{Definition of addition}
  Fix \(a\in\N\).  We define \(a+b\) by recursion on \(b\):
  \[
    \boxed{a+0=a},
    \qquad
    \boxed{a+\suc b=\suc{a+b}}.
  \]

  Thus the second input tells us how many times to take the successor of \(a\).
  For example,
  \begin{align*}
    a+\suc{\suc 0}
      &=\suc{a+\suc 0}\\
      &=\suc{\suc{a+0}}\\
      &=\suc{\suc a}.
  \end{align*}

  Since \(1=\suc 0\), the defining equations immediately give
  \[
    a+1=a+\suc 0=\suc{a+0}=\suc a.
  \]
\end{frame}

\begin{frame}{Zero on the left}
  \begin{block}{Proposition}
    For every \(a\in\N\), \(0+a=a\).
  \end{block}

  \begin{proof}
    We use induction on \(a\).

    \emph{Base case.} By the first defining equation,
    \(0+0=0\).

    \emph{Inductive step.} Suppose that \(0+n=n\).  Then
    \[
      0+\suc n
        =\suc{0+n}
        =\suc n.
    \]
    The first equality is the second defining equation for addition; the
    second uses the induction hypothesis.  Hence the statement passes from
    \(n\) to \(\suc n\).

    By induction, \(0+a=a\) for all \(a\in\N\).
  \end{proof}
\end{frame}

\begin{frame}{Moving a successor through a sum}
  \begin{block}{Proposition}
    For all \(a,b\in\N\),
    \[
      \suc a+b=\suc{a+b}.
    \]
  \end{block}

  \begin{proof}
    Fix \(a\) and use induction on \(b\).

    \emph{Base case.}
    \(\suc a+0=\suc a=\suc{a+0}\).

    \emph{Inductive step.} Assume \(\suc a+n=\suc{a+n}\).  Then
    \begin{align*}
      \suc a+\suc n
        &=\suc{\suc a+n} &&\text{by the definition of addition}\\
        &=\suc{\suc{a+n}} &&\text{by the induction hypothesis}\\
        &=\suc{a+\suc n} &&\text{by the definition of addition}.
    \end{align*}
    This proves the required identity for \(\suc n\).
  \end{proof}
\end{frame}

\begin{frame}{Associativity of addition}
  \footnotesize
  \begin{block}{Theorem}
    For all \(a,b,c\in\N\),
    \[
      (a+b)+c=a+(b+c).
    \]
  \end{block}

  \begin{proof}
    Fix \(a,b\) and use induction on \(c\).

    \emph{Base case.}
    \((a+b)+0=a+b=a+(b+0)\).

    \emph{Inductive step.} Suppose \((a+b)+n=a+(b+n)\).  Then
    \begin{align*}
      (a+b)+\suc n
        &=\suc{(a+b)+n}\\
        &=\suc{a+(b+n)}\\
        &=a+\suc{b+n}\\
        &=a+(b+\suc n).
    \end{align*}
    Here the second equality is the induction hypothesis, and the other
    equalities are the defining equation for addition, read in either direction.
  \end{proof}
\end{frame}

\begin{frame}{Commutativity of addition}
  \begin{block}{Theorem}
    For all \(a,b\in\N\), \(a+b=b+a\).
  \end{block}

  \begin{proof}
    Fix \(b\) and use induction on \(a\).

    \emph{Base case.} We already proved \(0+b=b\), while \(b+0=b\) by
    definition.  Hence \(0+b=b+0\).

    \emph{Inductive step.} Suppose \(n+b=b+n\).  Using the formula for
    moving a successor through a sum,
    \begin{align*}
      \suc n+b
        &=\suc{n+b}\\
        &=\suc{b+n} &&\text{by the induction hypothesis}\\
        &=b+\suc n &&\text{by the definition of addition}.
    \end{align*}
    Therefore the equality holds for \(\suc n\), and induction completes the proof.
  \end{proof}
\end{frame}

\begin{frame}{When can a sum be zero?}
  \small
  \begin{block}{Proposition}
    If \(a+b=0\), then \(a=0\) and \(b=0\).
  \end{block}

  \begin{proof}
    First inspect \(b\).

    If \(b=0\), then \(a+b=a+0=a\), so the assumption gives \(a=0\).

    If \(b=\suc n\), then
    \[
      a+b=a+\suc n=\suc{a+n},
    \]
    which cannot equal \(0\).  Thus the second possibility is ruled out, and
    \(b=0\); the preceding paragraph then gives \(a=0\).
  \end{proof}

  Conversely, \(0+0=0\), so in fact
  \[
    a+b=0\quad\Longleftrightarrow\quad a=0\text{ and }b=0.
  \]
\end{frame}

\begin{frame}{Cancellation for addition}
  \begin{block}{Theorem}
    If \(a+b=a+c\), then \(b=c\).
  \end{block}

  \begin{proof}
    We use induction on \(a\), proving the assertion simultaneously for all
    \(b,c\in\N\).

    \emph{Base case.} If \(0+b=0+c\), then \(b=c\), because \(0+b=b\) and
    \(0+c=c\).

    \emph{Inductive step.} Assume cancellation is valid with \(n\) on the left.
    If
    \[
      \suc n+b=\suc n+c,
    \]
    moving the successors through the sums gives
    \(\suc{n+b}=\suc{n+c}\).  Uniqueness of successors yields \(n+b=n+c\),
    and the induction hypothesis now yields \(b=c\).
  \end{proof}
\end{frame}

\subsection{Multiplication}

\begin{frame}{Definition of multiplication}
  Fix \(a\in\N\).  We define \(ab\) by recursion on \(b\):
  \[
    \boxed{a\cdot0=0},
    \qquad
    \boxed{a\cdot\suc b=ab+a}.
  \]

  Multiplication is therefore repeated addition.  For example,
  \begin{align*}
    a\cdot\suc{\suc 0}
      &=a\cdot\suc 0+a\\
      &=(a\cdot0+a)+a\\
      &=(0+a)+a\\
      &=a+a.
  \end{align*}
\end{frame}

\begin{frame}{Moving a successor through a product}
  \footnotesize
  \begin{block}{Proposition}
    For all \(a,b\in\N\),
    \[
      \suc a\cdot b=ab+b.
    \]
  \end{block}

  \begin{proof}
    Fix \(a\) and use induction on \(b\).

    \emph{Base case.}
    \(\suc a\cdot0=0=a\cdot0+0\).

    \emph{Inductive step.} Suppose \(\suc a\cdot n=an+n\).  Then
    \begin{align*}
      \suc a\cdot\suc n
        &=\suc a\cdot n+\suc a\\
        &=(an+n)+\suc a\\
        &=\suc{(an+n)+a}\\
        &=\suc{(an+a)+n}\\
        &=a\cdot\suc n+\suc n.
    \end{align*}
    Associativity and commutativity of addition justify the fourth equality.
  \end{proof}
\end{frame}

\begin{frame}{Zero on the left of a product}
  \begin{block}{Proposition}
    For every \(a\in\N\), \(0\cdot a=0\).
  \end{block}

  \begin{proof}
    We use induction on \(a\).

    \emph{Base case.} \(0\cdot0=0\) by definition.

    \emph{Inductive step.} Suppose \(0\cdot n=0\).  Then
    \[
      0\cdot\suc n=0\cdot n+0=0+0=0.
    \]
    Thus the property passes to the successor, and induction proves it for
    every \(a\).
  \end{proof}

  Together with the defining identity \(a\cdot0=0\), this says that \(0\) is
  absorbing on both sides of multiplication.
\end{frame}

\begin{frame}{Left distributivity}
  \footnotesize
  \begin{block}{Theorem}
    For all \(a,b,c\in\N\),
    \[
      a(b+c)=ab+ac.
    \]
  \end{block}

  \begin{proof}
    Fix \(a,b\) and use induction on \(c\).

    \emph{Base case.}
    \(a(b+0)=ab=ab+0=ab+a\cdot0\).

    \emph{Inductive step.} Suppose \(a(b+n)=ab+an\).  Then
    \begin{align*}
      a(b+\suc n)
        &=a\suc{(b+n)}\\
        &=a(b+n)+a\\
        &=(ab+an)+a\\
        &=ab+(an+a)\\
        &=ab+a\suc n.
    \end{align*}
    We used the recursion equations and associativity of addition.
  \end{proof}
\end{frame}

\begin{frame}{The multiplicative identity}
  \begin{block}{Proposition}
    For every \(a\in\N\), \(a\cdot1=a\).
  \end{block}

  Since \(1=\suc 0\), the definition of multiplication gives
  \[
    a\cdot1=a\cdot\suc 0=a\cdot0+a=0+a=a.
  \]

  \begin{block}{A useful observation}
    The equality \(1\cdot a=a\) will follow once multiplication is known to be
    commutative.  It can also be proved directly by induction on \(a\).

    The base case is \(1\cdot0=0\).  If \(1\cdot n=n\), then
    \[
      1\cdot\suc n=1\cdot n+1=n+1=\suc n.
    \]
  \end{block}
\end{frame}

\begin{frame}{Commutativity of multiplication}
  \begin{block}{Theorem}
    For all \(a,b\in\N\), \(ab=ba\).
  \end{block}

  \begin{proof}
    Fix \(a\) and use induction on \(b\).

    \emph{Base case.}
    \(a\cdot0=0=0\cdot a\).

    \emph{Inductive step.} Suppose \(an=na\).  Then
    \begin{align*}
      a\cdot\suc n
        &=an+a &&\text{by the definition of multiplication}\\
        &=na+a &&\text{by the induction hypothesis}\\
        &=\suc n\cdot a &&\text{by the successor--product formula}.
    \end{align*}
    Hence \(a\cdot b=b\cdot a\) for all \(b\).
  \end{proof}
\end{frame}

\begin{frame}{Right distributivity}
  \begin{block}{Theorem}
    For all \(a,b,c\in\N\),
    \[
      (a+b)c=ac+bc.
    \]
  \end{block}

  \begin{proof}
    We convert right multiplication into left multiplication, use left
    distributivity, and convert back:
    \begin{align*}
      (a+b)c
        &=c(a+b) &&\text{commutativity of multiplication}\\
        &=ca+cb &&\text{left distributivity}\\
        &=ac+bc &&\text{commutativity of multiplication}.
    \end{align*}
  \end{proof}

  This short argument illustrates an important strategy: a symmetry theorem can
  turn a new statement into one already proved.
\end{frame}

\begin{frame}{Associativity of multiplication}
  \small
  \begin{block}{Theorem}
    For all \(a,b,c\in\N\),
    \[
      (ab)c=a(bc).
    \]
  \end{block}

  \begin{proof}
    Fix \(a,b\) and use induction on \(c\).

    \emph{Base case.}
    \((ab)0=0=a0=a(b0)\).

    \emph{Inductive step.} Suppose \((ab)n=a(bn)\).  Then
    \begin{align*}
      (ab)\suc n
        &=(ab)n+ab\\
        &=a(bn)+ab\\
        &=a(bn+b)\\
        &=a(b\suc n).
    \end{align*}
    The third equality uses left distributivity in reverse.
  \end{proof}
\end{frame}

\begin{frame}{No zero divisors}
  \small
  \begin{block}{Theorem}
    If \(a\ne0\) and \(b\ne0\), then \(ab\ne0\).
  \end{block}

  Since every nonzero natural number is a successor, write
  \(a=\suc p=p+1\) and \(b=\suc q=q+1\).  Then
  \begin{align*}
    ab
      &=(p+1)(q+1)\\
      &=p(q+1)+1(q+1)\\
      &=(pq+p)+(q+1)\\
      &=(pq+p+q)+1\\
      &=\suc{pq+p+q}.
  \end{align*}
  The final expression is a successor, so it cannot be \(0\).  Therefore
  \(ab\ne0\).

  \begin{alertblock}{Equivalent form}
    If \(ab=0\), then \(a=0\) or \(b=0\).
  \end{alertblock}
\end{frame}

\begin{frame}{The predecessor function}
  Define \(\operatorname{pred}\colon\N\to\N\) by
  \[
    \operatorname{pred}(0)=0,
    \qquad
    \operatorname{pred}(\suc a)=a.
  \]

  Consequently,
  \[
    \operatorname{pred}(a+1)=a
  \]
  for every \(a\), because \(a+1=\suc a\).

  Conversely, if \(a\ne0\), then \(a\) is a successor, say \(a=\suc n\), and
  \[
    \operatorname{pred}(a)+1
      =\operatorname{pred}(\suc n)+1
      =n+1
      =\suc n
      =a.
  \]

  The condition \(a\ne0\) is essential: our definition gives
  \(\operatorname{pred}(0)+1=1\ne0\).
\end{frame}

\begin{frame}{The algebra obtained so far}
  For all \(a,b,c\in\N\), we have established
  \begin{align*}
    a+b&=b+a, & (a+b)+c&=a+(b+c), & a+0&=a,\\
    ab&=ba, & (ab)c&=a(bc), & a\cdot1&=a,\\
    a(b+c)&=ab+ac, & (a+b)c&=ac+bc, & a\cdot0&=0.
  \end{align*}

  Thus \(\N\), together with \(0,1,+,\cdot\), is a
  \alert{commutative semiring}.  In addition,
  \begin{itemize}
    \item \(0\ne1\);
    \item addition admits cancellation;
    \item a product of two nonzero numbers is nonzero.
  \end{itemize}

  We now build the order directly from addition.
\end{frame}

\subsection{The order relation}

\begin{frame}{Definition of \(\leq\)}
  \begin{block}{Definition}
    For \(a,b\in\N\), we write
    \[
      \boxed{a\le b}
      \qquad\text{if and only if}\qquad
      \boxed{\text{there is an }x\in\N\text{ such that }b=a+x.}
    \]
  \end{block}

  The number \(x\) is the \alert{gap} from \(a\) to \(b\).

  Immediate examples are
  \begin{itemize}
    \item \(0\le a\), because \(a=0+a\);
    \item \(a\le a\), because \(a=a+0\);
    \item \(a\le\suc a\), because \(\suc a=a+1\);
    \item \(a\le a+b\), because \(a+b=a+b\).
  \end{itemize}

  We must prove that this relation really has all the expected order properties.
\end{frame}

\begin{frame}{Reflexivity of \(\leq\)}
  \begin{block}{Proposition}
    For every \(a\in\N\),
    \[
      a\le a.
    \]
  \end{block}

  \begin{proof}
    By definition, we must find an \(x\in\N\) such that
    \[
      a=a+x.
    \]
    Choose \(x=0\).  The additive identity gives
    \[
      a=a+0,
    \]
    so \(0\) is a gap from \(a\) to itself.  Hence \(a\le a\).
  \end{proof}
\end{frame}

\begin{frame}{Transitivity of \(\leq\)}
  \begin{block}{Proposition}
    If \(a\le b\) and \(b\le c\), then \(a\le c\).
  \end{block}

  \begin{proof}
    Since \(a\le b\), choose \(x\in\N\) such that
    \[
      b=a+x.
    \]
    Since \(b\le c\), choose \(y\in\N\) such that
    \[
      c=b+y.
    \]
    Substituting the first equality into the second and using associativity,
    \[
      c=(a+x)+y=a+(x+y).
    \]
    Because \(x+y\in\N\), it is a gap from \(a\) to \(c\).  Hence \(a\le c\).
  \end{proof}
\end{frame}

\begin{frame}{Antisymmetry of \(\leq\): setting up the gaps}
  \begin{block}{Theorem}
    If \(a\le b\) and \(b\le a\), then \(a=b\).
  \end{block}

  Choose gaps \(x,y\in\N\) such that
  \[
    b=a+x,
    \qquad
    a=b+y.
  \]
  Substitution gives
  \[
    a=b+y=(a+x)+y=a+(x+y).
  \]
  But also \(a=a+0\).  Therefore
  \[
    a+0=a+(x+y).
  \]
  Cancellation for addition yields
  \[
    0=x+y.
  \]
\end{frame}

\begin{frame}{Antisymmetry of \(\leq\): finishing the proof}
  From \(x+y=0\), the zero-sum proposition gives \(x=0\) and \(y=0\).
  In particular, the first gap vanishes, and therefore
  \[
    b=a+x=a+0=a.
  \]
  Hence \(a=b\), as required.

\bigskip

  Notice the chain of earlier results used here:
  \[
    \text{associativity}
    \;\Longrightarrow\;
    \text{addition cancellation}
    \;\Longrightarrow\;
    \text{antisymmetry}.
  \]
\end{frame}

\begin{frame}{Totality of \(\leq\): the induction}
  \begin{block}{Theorem}
    For every \(a,b\in\N\), either \(a\le b\) or \(b\le a\).
  \end{block}

  Fix \(b\) and use induction on \(a\).

  \emph{Base case.} For \(a=0\), we have \(0\le b\), witnessed by the gap \(b\).

  \emph{Inductive step.} Suppose that either \(n\le b\) or \(b\le n\).
  We must compare \(\suc n\) with \(b\).

  If \(b\le n\), then \(b\le n\le\suc n\), so transitivity gives
  \(b\le\suc n\).

  It remains to study the case \(n\le b\).
\end{frame}

\begin{frame}{Totality of \(\leq\): analysing the remaining gap}
  Suppose \(n\le b\).  Choose \(x\in\N\) with \(b=n+x\).
  There are two possibilities.

  \begin{itemize}
    \item If \(x=0\), then \(b=n\), and hence \(b=n\le\suc n\).
    \item If \(x=\suc y\), then
      \begin{align*}
        b=n+\suc y
          &=\suc{n+y}\\
          &=\suc n+y.
      \end{align*}
      Thus \(y\) is a gap from \(\suc n\) to \(b\), so \(\suc n\le b\).
  \end{itemize}

  In every case, \(\suc n\le b\) or \(b\le\suc n\).  The inductive step is
  complete, and therefore the order is total.
\end{frame}

\begin{frame}{A useful successor test}
  \begin{block}{Proposition}
    For all \(a,b\in\N\),
    \[
      a\le\suc b
      \quad\Longleftrightarrow\quad
      a=\suc b\ \text{ or }\ a\le b.
    \]
  \end{block}

  Suppose first that \(\suc b=a+x\).
  \begin{itemize}
    \item If \(x=0\), then \(\suc b=a\), so \(a=\suc b\).
    \item If \(x=\suc y\), then
      \(\suc b=a+\suc y=\suc{a+y}\).  Uniqueness of successors gives
      \(b=a+y\), hence \(a\le b\).
  \end{itemize}

  Conversely, equality gives \(a\le\suc b\) by reflexivity; and if \(a\le b\),
  then \(a\le b\le\suc b\) by transitivity.
\end{frame}

\begin{frame}{We have a linear order}
  The relation \(\leq\) is
  \begin{itemize}
    \item \alert{reflexive}: \(a\le a\);
    \item \alert{transitive}: \(a\le b\le c\) implies \(a\le c\);
    \item \alert{antisymmetric}: \(a\le b\) and \(b\le a\) imply \(a=b\);
    \item \alert{total}: either \(a\le b\) or \(b\le a\).
  \end{itemize}

  Therefore \((\N,\le)\) is a \alert{linearly ordered set}.

  The gap definition also gives, without further work,
  \[
    a\le a+b
    \qquad\text{and}\qquad
    a\le b+a,
  \]
  because addition is commutative.
\end{frame}

\subsection{Strict order and arithmetic}

\begin{frame}{Definition of strict inequality}
  \begin{block}{Definition}
    For \(a,b\in\N\),
    \[
      a<b
      \quad\Longleftrightarrow\quad
      a\le b\ \text{ and }\ b\nleq a.
    \]
  \end{block}

  \begin{block}{Equivalent form}

    \[
      a<b
      \quad\Longleftrightarrow\quad
      a\le b\ \text{ and }\ a\ne b.
    \]
  \end{block}

  Indeed, if \(a<b\) and \(a=b\), reflexivity would give \(b\le a\), a
  contradiction.  Conversely, if \(a\le b\), \(a\ne b\), and \(b\le a\),
  antisymmetry would force \(a=b\), again a contradiction.
\end{frame}

\begin{frame}{Positive numbers}
  \begin{block}{Proposition}
    For every \(a\in\N\),
    \[
      a\ne0\quad\Longleftrightarrow\quad 0<a.
    \]
  \end{block}

  \begin{proof}
    We always have \(0\le a\).  If \(a\ne0\), the equivalent form of strict
    inequality therefore gives \(0<a\).

    Conversely, if \(0<a\), then strict inequality implies \(0\ne a\), which
    is the same as \(a\ne0\).
  \end{proof}

  Thus, for natural numbers, \alert{nonzero} and \alert{positive} mean exactly
  the same thing.
\end{frame}

\begin{frame}{Strict inequality means a positive gap}
  \begin{block}{Theorem}
    For \(a,b\in\N\),
    \[
      a<b
      \quad\Longleftrightarrow\quad
      \text{there is an }x\ne0\text{ such that }b=a+x.
    \]
  \end{block}

  If \(a<b\), choose \(x\) with \(b=a+x\).  If \(x=0\), then
  \(b=a+0=a\), contradicting \(a\ne b\).  Hence \(x\ne0\).

  Conversely, suppose \(b=a+x\) with \(x\ne0\).  Then \(a\le b\).  If \(a=b\),
  we would have
  \[
    a+0=a=b=a+x.
  \]
  Cancellation would give \(x=0\), a contradiction.  Thus \(a\ne b\), and
  consequently \(a<b\).
\end{frame}

\begin{frame}{Addition preserves order}
  \begin{block}{Theorem}
    If \(a\le b\), then \(a+c\le b+c\).
  \end{block}

  Choose \(x\in\N\) with \(b=a+x\).  We look for a gap from \(a+c\) to \(b+c\).
  Using associativity and commutativity,
  \begin{align*}
    b+c
      &=(a+x)+c\\
      &=a+(x+c)\\
      &=a+(c+x)\\
      &=(a+c)+x.
  \end{align*}
  Thus the same \(x\) is a gap from \(a+c\) to \(b+c\), proving
  \(a+c\le b+c\).

  By commutativity, adding \(c\) on the left also preserves order.
\end{frame}

\begin{frame}{Cancelling the same summand from an inequality}
  \begin{block}{Theorem}
    If \(a+b\le a+c\), then \(b\le c\).
  \end{block}

  Choose \(x\in\N\) such that
  \[
    a+c=(a+b)+x.
  \]
  By associativity, the right-hand side is \(a+(b+x)\), so
  \[
    a+c=a+(b+x).
  \]
  Addition cancellation gives
  \[
    c=b+x.
  \]
  Hence \(x\) is a gap from \(b\) to \(c\), and \(b\le c\).

  Together with order preservation, this yields
  \[
    a+b\le a+c\quad\Longleftrightarrow\quad b\le c.
  \]
\end{frame}

\begin{frame}{Positive multiplication preserves strict order}
  \begin{block}{Theorem}
    If \(0<c\) and \(a<b\), then \(ca<cb\).
  \end{block}

  Because \(a<b\), write \(b=a+x\) with \(x\ne0\).  Because \(0<c\), we also
  have \(c\ne0\).  Distributivity gives
  \[
    cb=c(a+x)=ca+cx.
  \]
  Since \(c\ne0\) and \(x\ne0\), the no-zero-divisors theorem gives
  \(cx\ne0\).  Thus \(cb\) is obtained from \(ca\) by adding the nonzero gap
  \(cx\).  The positive-gap characterization now gives
  \[
    ca<cb.
  \]
\end{frame}

\begin{frame}{Multiplication on the right}
  \begin{block}{Corollary}
    If \(0<c\) and \(a<b\), then \(ac<bc\).
  \end{block}

  By commutativity of multiplication,
  \[
    ac=ca<cb=bc.
  \]

  \begin{alertblock}{Why positivity is necessary}
    If \(c=0\), then \(ac=0=bc\), even when \(a<b\).  Multiplication by zero
    destroys strict inequality.
  \end{alertblock}

  For weak inequalities, no positivity assumption is needed: if \(a\le b\),
  write \(b=a+x\); then \(bc=ac+xc\), so \(ac\le bc\).
\end{frame}

\begin{frame}{Cancellation for multiplication: reduction to order}
  \begin{block}{Theorem}
    Suppose \(a\ne0\).  If \(ab=ac\), then \(b=c\).
  \end{block}

  Since the order is total, either \(b\le c\) or \(c\le b\).

  Assume first that \(b\le c\).  If \(b\ne c\), then \(b<c\).  Since
  \(a\ne0\) is equivalent to \(0<a\), strict order preservation gives
  \[
    ab<ac,
  \]
  contradicting \(ab=ac\).  Therefore \(b=c\).

  The case \(c\le b\) is symmetric: if \(c\ne b\), then \(c<b\), hence
  \(ac<ab\), again contradicting the assumed equality.
\end{frame}

\begin{frame}{Right cancellation for multiplication}
  \begin{block}{Corollary}
    Suppose \(a\ne0\).  If \(ba=ca\), then \(b=c\).
  \end{block}

  Commutativity converts the hypothesis into
  \[
    ab=ac.
  \]
  The multiplication cancellation theorem therefore gives \(b=c\).

  \begin{alertblock}{Why \(a\ne0\) is necessary}
    For \(a=0\), the equality \(b0=c0\) holds for every \(b,c\in\N\), so no
    conclusion about \(b\) and \(c\) is possible.
  \end{alertblock}
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  Starting from the natural-number principles, we proved that
  \begin{itemize}
    \item addition and multiplication obey the usual associative,
      commutative, distributive, and identity laws;
    \item sums and products have strong cancellation properties;
    \item nonzero products are nonzero;
    \item the relation \(a\le b\iff b=a+x\) is a linear order;
    \item \(a<b\) means exactly that the gap from \(a\) to \(b\) is positive;
    \item addition preserves order, and multiplication by a positive number
      preserves strict order.
  \end{itemize}

\end{frame}

\begin{frame}{An equation without solutions in \(\N\)}
  Suppose that some \(x\in\N\) satisfied
  \[
    3+x=1.
  \]
  By commutativity and the recursive definition of addition,
  \[
    3+x=x+3=\suc{\suc{\suc x}},
    \qquad
    1=\suc 0.
  \]
  Hence
  \[
    \suc{\suc{\suc x}}=\suc 0.
  \]
  Injectivity of the successor operation would then give
  \[
    \suc{\suc x}=0,
  \]
  which is impossible because zero is not a successor.  Therefore
  \(3+x=1\) has no solution in \(\N\).
\end{frame}

\section{The integers}

\subsection{Formal differences}

\begin{frame}{Why introduce the integers?}
  In \(\N\), the equation
  \[
    3+x=1
  \]
  has no solution.  We want a larger number system in which it has a solution
  (namely \(x=-2\)).

  Every integer can be thought of as a difference
  \[
    a-b,\qquad a,b\in\N.
  \]
  Since subtraction is not yet available, we begin with the ordered pair
  \((a,b)\), intended to represent that difference.

  \begin{alertblock}{The central difficulty}
    The same integer has many representations:
    \[
      5-2=4-1=3-0.
    \]
    We must decide exactly when two pairs represent the same difference.
  \end{alertblock}
\end{frame}

\begin{frame}{The pre-integers}
  \begin{block}{Definition}
    The set of \alert{pre-integers} is
    \[
      \Z_{\mathrm{pre}}=\N^2,
    \]
    the set of ordered pairs \((a,b)\) of natural numbers.
  \end{block}

  The pair \((a,b)\) records the difference we intend to create:
  \[
    (a,b)\quad\longleftrightarrow\quad a-b.
  \]

  Every pair is allowed: a difference makes sense for any \(a,b\in\N\).
  The only problem, addressed next, is that different pre-integers can
  represent the same difference.
\end{frame}

\begin{frame}{Equality of formal differences}
  \begin{block}{Definition}
    For pre-integers \((a,b),(c,d)\in\Z_{\mathrm{pre}}\), declare
    \[
      (a,b)\sim(c,d)
      \quad\Longleftrightarrow\quad
      a+d=b+c.
    \]
  \end{block}

  The equation uses only addition in \(\N\), but it expresses
  \[
    a-b=c-d.
  \]
  Indeed, cross-adding \(b+d\) to the informal equation gives precisely
  \(a+d=b+c\).

  For example,
  \[
    (5,2)\sim(4,1)
    \qquad\text{because}\qquad
    5+1=2+4.
  \]
\end{frame}

\begin{frame}{Reflexivity and symmetry}
  \begin{block}{Proposition}
    The relation \(\sim\) is reflexive and symmetric.
  \end{block}

  \begin{proof}
    For reflexivity, let \((a,b)\in\Z_{\mathrm{pre}}\).  The equality \(a+b=b+a\)
    shows directly that \((a,b)\sim(a,b)\).

    For symmetry, suppose \((a,b)\sim(c,d)\), so \(a+d=b+c\).  Then
    \[
      c+b=b+c=a+d=d+a.
    \]
    The first and last equalities use commutativity.  Thus
    \(c+b=d+a\), which is exactly \((c,d)\sim(a,b)\).
  \end{proof}
\end{frame}

\begin{frame}{Transitivity}
  \small
  \begin{block}{Proposition}
    If \((a,b)\sim(c,d)\) and \((c,d)\sim(e,f)\), then
    \((a,b)\sim(e,f)\).
  \end{block}

  \begin{proof}
    The assumptions say
    \[
      a+d=b+c,
      \qquad
      c+f=d+e.
    \]
    Add the two equalities and rearrange:
    \[
      (a+f)+(c+d)=(b+e)+(c+d).
    \]
    Addition in \(\N\) admits cancellation, so cancelling the common summand
    \(c+d\) yields
    \[
      a+f=b+e.
    \]
    This is precisely the condition \((a,b)\sim(e,f)\).
  \end{proof}
\end{frame}

\begin{frame}{The integers}
  Reflexivity, symmetry, and transitivity show that \(\sim\) is an equivalence
  relation on the pre-integers.

  \begin{block}{Definition}
    The set of integers is the set of equivalence classes
    \[
      \boxed{\Z=\Z_{\mathrm{pre}}/{\sim}}.
    \]
    We write
    \[
      \cl{a}{b}
    \]
    for the class represented by \((a,b)\).
  \end{block}

  Thus the equality criterion in \(\Z\) is
  \[
    \boxed{\cl{a}{b}=\cl{c}{d}
      \quad\Longleftrightarrow\quad a+d=b+c.}
  \]
  This criterion will be used repeatedly.
\end{frame}

\begin{frame}{The universal property of a quotient}
  \footnotesize
  Let \(X\) be a set with an equivalence relation \(\sim\), let \(Y\) be a
  set, and write \(q\colon X\to X/{\sim}\), \(q(x)=[x]\), for the quotient map.

  \begin{block}{Universal property}
    A function \(f\colon X\to Y\) descends to a function
    \(\bar f\colon X/{\sim}\to Y\) satisfying
    \[
      \bar f([x])=f(x)
    \]
    if and only if it is constant on equivalence classes:
    \[
      x\sim x'\quad\Longrightarrow\quad f(x)=f(x').
    \]
    When it exists, \(\bar f\) is unique; equivalently,
    \(f=\bar f\circ q\).
  \end{block}

  Thus functions \(X/{\sim}\to Y\) correspond exactly to functions
  \(X\to Y\) that are constant on equivalence classes.

  \begin{alertblock}{How to define a function on a quotient}
    Give a formula on representatives and prove that equivalent representatives
    have the same image.  The descended function is then uniquely determined.
  \end{alertblock}
\end{frame}

\subsection{Arithmetic on integers}

\begin{frame}{Negation and addition}
  The familiar rules for differences suggest the definitions
  \[
    -(a-b)=b-a,
    \qquad
    (a-b)+(c-d)=(a+c)-(b+d).
  \]

  \begin{block}{Definitions}
    For integer classes, set
    \[
      -\cl{a}{b}=\cl{b}{a},
      \qquad
      \cl{a}{b}+\cl{c}{d}=\cl{a+c}{b+d}.
    \]
  \end{block}

  Before using these formulas, we must prove that their values do not depend
  on the chosen representatives.
\end{frame}

\begin{frame}{Negation is well defined}
  \begin{block}{Proposition}
    If \((a,b)\sim(c,d)\), then \((b,a)\sim(d,c)\).
  \end{block}

  \begin{proof}
    The assumption is
    \[
      a+d=b+c.
    \]
    To prove \((b,a)\sim(d,c)\), we need
    \[
      b+c=a+d,
    \]
    which is the same equality read in the opposite direction.
  \end{proof}

  Hence equivalent pairs have equivalent negatives, so
  \(-\cl{a}{b}\) is independent of the representative \((a,b)\).
\end{frame}

\begin{frame}{Addition is well defined}
  \small
  Suppose
  \[
    (a,b)\sim(a',b'),
    \qquad
    (c,d)\sim(c',d').
  \]
  Thus \(a+b'=b+a'\) and \(c+d'=d+c'\).  Adding these equations and
  rearranging gives
  \[
    (a+c)+(b'+d')=(b+d)+(a'+c').
  \]
  Therefore
  \[
    (a+c,b+d)\sim(a'+c',b'+d').
  \]

  We have proved that replacing either summand by an equivalent pair does not
  alter the class of the sum.  Consequently,
  \[
    \cl{a}{b}+\cl{c}{d}=\cl{a+c}{b+d}
  \]
  is a well-defined operation on \(\Z\).
\end{frame}

\begin{frame}{Discovering multiplication}
  Formally expanding a product of differences gives
  \[
    (a-b)(c-d)=(ac+bd)-(ad+bc).
  \]
  This suggests the following definition.

  \begin{block}{Definition}
    \[
      \boxed{
      \cl{a}{b}\cl{c}{d}
        =\cl{ac+bd}{ad+bc}.}
    \]
  \end{block}

  For example,
  \[
    \cl{5}{2}\cl{1}{4}
      =\cl{5\cdot1+2\cdot4}{5\cdot4+2\cdot1}
      =\cl{13}{22},
  \]
  which represents \(-9\), as expected from \(3\cdot(-3)=-9\).
\end{frame}

\begin{frame}{Multiplication respects one representative}
  \footnotesize
  Suppose \((a,b)\sim(c,d)\), so \(a+d=b+c\), and fix \((e,f)\).
  Multiplying the given equality by \(e\) and by \(f\) gives
  \[
    ae+de=be+ce,
    \qquad
    af+df=bf+cf.
  \]
  We must compare
  \[
    (ae+bf,af+be)
    \quad\text{and}\quad
    (ce+df,cf+de).
  \]
  Their cross-sums satisfy
  \begin{align*}
    (ae+bf)+(cf+de)
      &=(ae+de)+(bf+cf)\\
      &=(be+ce)+(af+df)\\
      &=(af+be)+(ce+df).
  \end{align*}
  Hence the two product pairs are equivalent.
\end{frame}

\begin{frame}{Multiplication is well defined}
  Pair multiplication is commutative:
  \[
    (ae+bf,af+be)=(ea+fb,eb+fa),
  \]
  by commutativity in \(\N\).

  The preceding argument therefore works in either factor.  If
  \[
    (a,b)\sim(a',b'),
    \qquad
    (c,d)\sim(c',d'),
  \]
  then
  \[
    (a,b)(c,d)\sim(a',b')(c,d)
      \sim(a',b')(c',d').
  \]
  Transitivity now shows that the product class is unchanged when both
  representatives are replaced.

  Thus multiplication is a well-defined operation on \(\Z\).
\end{frame}

\begin{frame}{Zero and one}
  Define
  \[
    0=\cl{0}{0},
    \qquad
    1=\cl{1}{0}.
  \]

  The equality criterion immediately gives a useful test:
  \[
    \boxed{\cl{a}{b}=0\quad\Longleftrightarrow\quad a=b.}
  \]
  Indeed,
  \[
    \cl{a}{b}=\cl{0}{0}
    \quad\Longleftrightarrow\quad
    a+0=b+0
    \quad\Longleftrightarrow\quad
    a=b.
  \]

  In particular \(0\ne1\), since \(0\ne1\) in \(\N\).
\end{frame}

\begin{frame}{Associativity of addition}
  \small
  \begin{block}{Theorem}
    For all \(x,y,z\in\Z\), \((x+y)+z=x+(y+z)\).
  \end{block}

  \begin{proof}
    Choose representatives
    \[
      x=\cl{a}{b},\qquad y=\cl{c}{d},\qquad z=\cl{e}{f}.
    \]
    Then
    \begin{align*}
      (x+y)+z
        &=\cl{(a+c)+e}{(b+d)+f},\\
      x+(y+z)
        &=\cl{a+(c+e)}{b+(d+f)}.
    \end{align*}
    The two displayed pairs are equal by associativity of addition in \(\N\).
    Therefore their integer classes are equal.
  \end{proof}
\end{frame}

\begin{frame}{The additive identity and additive inverses}
  Let \(x=\cl{a}{b}\).

  \begin{block}{Zero is an identity}
    \[
      x+0=\cl{a+0}{b+0}=\cl{a}{b}=x.
    \]
    Commutativity of natural addition similarly gives \(0+x=x\).
  \end{block}

  \begin{block}{Every integer has an additive inverse}
    \[
      x+(-x)=\cl{a+b}{b+a}.
    \]
    Since \(a+b=b+a\), the zero criterion gives
    \[
      x+(-x)=0.
    \]
  \end{block}
\end{frame}

\begin{frame}{Commutativity and subtraction}
  If \(x=\cl{a}{b}\) and \(y=\cl{c}{d}\), then
  \[
    x+y=\cl{a+c}{b+d}
       =\cl{c+a}{d+b}
       =y+x.
  \]
  Thus addition on \(\Z\) is commutative.

  \begin{block}{Definition of subtraction}
    We may now define
    \[
      x-y=x+(-y).
    \]
    In terms of representatives,
    \[
      \cl{a}{b}-\cl{c}{d}
        =\cl{a+d}{b+c}.
    \]
  \end{block}

  The operation that motivated the construction has now become available.
\end{frame}

\begin{frame}{The multiplicative identity and commutativity}
  Let \(x=\cl{a}{b}\).  Then
  \[
    1x
      =\cl{1}{0}\cl{a}{b}
      =\cl{1a+0b}{1b+0a}
      =\cl{a}{b}
      =x.
  \]

  If \(y=\cl{c}{d}\), then
  \begin{align*}
    xy&=\cl{ac+bd}{ad+bc},\\
    yx&=\cl{ca+db}{cb+da}.
  \end{align*}
  Corresponding coordinates are equal by commutativity in \(\N\), so \(xy=yx\).
  Consequently \(x1=1x=x\) as well.
\end{frame}

\begin{frame}{Associativity of multiplication: first coordinates}
  \footnotesize
  Let
  \[
    x=\cl{a}{b},\qquad y=\cl{c}{d},\qquad z=\cl{e}{f}.
  \]
  The first coordinate of a representative of \((xy)z\) is
  \begin{align*}
    (ac+bd)e+(ad+bc)f
      &=ace+bde+adf+bcf.
  \end{align*}
  The first coordinate of a representative of \(x(yz)\) is
  \begin{align*}
    a(ce+df)+b(cf+de)
      &=ace+adf+bcf+bde.
  \end{align*}
  These are equal by associativity, commutativity, and distributivity in \(\N\).
\end{frame}

\begin{frame}{Associativity of multiplication: second coordinates}
  \footnotesize
  With the same notation, the second coordinate of \((xy)z\) is
  \[
    (ac+bd)f+(ad+bc)e
      =acf+bdf+ade+bce.
  \]
  The second coordinate of \(x(yz)\) is
  \[
    a(cf+de)+b(ce+df)
      =acf+ade+bce+bdf.
  \]
  Again the two expressions are equal after rearranging their summands.
  Therefore the representing pairs are equal, and
  \[
    \boxed{(xy)z=x(yz).}
  \]
\end{frame}

\begin{frame}{Distributivity}
  \footnotesize
  Let \(x=\cl{a}{b}\), \(y=\cl{c}{d}\), and \(z=\cl{e}{f}\).
  The first coordinate of \(x(y+z)\) is
  \[
    a(c+e)+b(d+f)=ac+ae+bd+bf,
  \]
  while that of \(xy+xz\) is
  \[
    (ac+bd)+(ae+bf).
  \]
  They are equal after rearrangement.  The second coordinates are likewise
  \[
    a(d+f)+b(c+e)=ad+af+bc+be
  \]
  and
  \[
    (ad+bc)+(af+be).
  \]
  Hence \(x(y+z)=xy+xz\).  Commutativity of multiplication gives
  \((x+y)z=xz+yz\).
\end{frame}

\begin{frame}{The algebraic structure of \(\Z\)}
  We have proved that
  \begin{itemize}
    \item addition is associative and commutative, with identity \(0\);
    \item every integer has an additive inverse;
    \item multiplication is associative and commutative, with identity \(1\);
    \item multiplication distributes over addition;
    \item \(0\ne1\).
  \end{itemize}

  Therefore
  \[
    (\Z,0,1,+,-,\cdot)
  \]
  is a \alert{commutative ring}.

  Every one of these laws ultimately follows from the arithmetic of \(\N\) and
  the equality criterion for integer classes.
\end{frame}

\begin{frame}{The rule of signs}
  The classes \(\cl{n}{0}\) and \(\cl{0}{n}=-\cl{n}{0}\) represent a
  natural number and its negative.  Applying the multiplication formula
  directly gives, for \(m,n\in\N\),
  \[
  \begin{array}{c|cc}
    \cdot & \cl{n}{0} & \cl{0}{n}\\ \hline
    \cl{m}{0} & \cl{mn}{0} & \cl{0}{mn}\\[0.4em]
    \cl{0}{m} & \cl{0}{mn} & \cl{mn}{0}
  \end{array}
  \]

  This is the familiar rule of signs:
  \[
    (+)(+)=+,\quad (+)(-)=-,\quad (-)(+)=-,\quad (-)(-)=+.
  \]

  The table illustrates the signs of these particular representatives.
\end{frame}

\begin{frame}{A natural-number lemma}
  \small
  \begin{block}{Lemma}
    If \(a,b,c,d\in\N\) satisfy
    \[
      ad+bc=ac+bd,
    \]
    then \(a=b\) or \(c=d\).
  \end{block}

  The proof follows the induction used in the natural-number development.
  Induct on \(a\), allowing \(b\) to vary.

  If \(a=0\), the equality is \(bc=bd\), so either \(b=0=a\) or
  cancellation gives \(c=d\).  If \(a\) is a successor, split into the cases
  \(b=0\) and \(b=\suc f\).  The first again gives \(c=d\).  In the second,
  expanding both sides and cancelling the common terms \(c+d\) reduces the
  equality to the induction hypothesis.
\end{frame}

\begin{frame}{No zero divisors}
  \begin{block}{Theorem}
    If \(x\ne0\) and \(y\ne0\), then \(xy\ne0\).
  \end{block}

  \begin{proof}
    Choose arbitrary representatives
    \[
      x=\cl{a}{b},
      \qquad
      y=\cl{c}{d}.
    \]
    If \(xy=0\), the multiplication formula and the equality criterion give
    \[
      ad+bc=ac+bd.
    \]
    The natural-number lemma yields \(a=b\) or \(c=d\).  In the first case
    \(x=\cl{a}{a}=0\); in the second, \(y=\cl{c}{c}=0\).  Both conclusions
    contradict the hypotheses, so \(xy\ne0\).
  \end{proof}
\end{frame}

\begin{frame}{Cancellation for multiplication}
  \begin{block}{Theorem}
    If \(x\ne0\) and \(yx=zx\), then \(y=z\).
  \end{block}

  \begin{proof}
    Subtract \(zx\) from both sides:
    \[
      yx-zx=0.
    \]
    By distributivity,
    \[
      (y-z)x=0.
    \]
    If \(y-z\ne0\), then both factors would be nonzero, contradicting the
    no-zero-divisors theorem.  Hence \(y-z=0\), which is equivalent to \(y=z\).
  \end{proof}

  The assumption \(x\ne0\) is necessary, since \(y0=z0\) for all \(y,z\).
\end{frame}

\subsection{The natural numbers inside the integers}

\begin{frame}{The natural embedding}
  \begin{block}{Definition}
    Define
    \[
      \iota\colon\N\longrightarrow\Z,
      \qquad
      \iota(n)=\cl{n}{0}.
    \]
  \end{block}

  This map sends a natural number to the integer represented by \(n-0\).
  It preserves the distinguished elements:
  \[
    \iota(0)=\cl{0}{0}=0,
    \qquad
    \iota(1)=\cl{1}{0}=1.
  \]
\end{frame}

\begin{frame}{The embedding preserves arithmetic}
  For \(a,b\in\N\),
  \begin{align*}
    \iota(a)+\iota(b)
      &=\cl{a}{0}+\cl{b}{0}\\
      &=\cl{a+b}{0}
       =\iota(a+b),
  \end{align*}
  and
  \begin{align*}
    \iota(a)\iota(b)
      &=\cl{a}{0}\cl{b}{0}\\
      &=\cl{ab+0\cdot0}{a\cdot0+0\cdot b}\\
      &=\cl{ab}{0}
       =\iota(ab).
  \end{align*}

  Thus natural addition and multiplication agree with the integer operations.
\end{frame}

\begin{frame}{The embedding is injective}
  \begin{block}{Proposition}
    If \(\iota(a)=\iota(b)\), then \(a=b\).
  \end{block}

  \begin{proof}
    The equality
    \[
      \cl{a}{0}=\cl{b}{0}
    \]
    is equivalent to
    \[
      a+0=0+b.
    \]
    Hence \(a=b\).
  \end{proof}

  We may therefore identify \(\N\) with its image in \(\Z\), obtaining the
  familiar inclusion
  \[
    \N\subseteq\Z.
  \]
\end{frame}

\subsection{The order relation}

\begin{frame}{Order as a nonnegative gap}
  \begin{block}{Definition}
    For \(x,y\in\Z\), define
    \[
      \boxed{x\le y
      \quad\Longleftrightarrow\quad
      y=x+\iota(n)\text{ for some }n\in\N.}
    \]
  \end{block}

  Thus \(x\le y\) means that the gap \(y-x\) is a natural number.

  Immediate examples are
  \[
    x\le x
    \quad\text{using }n=0,
    \qquad
    0\le1
    \quad\text{using }n=1.
  \]
\end{frame}

\begin{frame}{Reflexivity of the integer order}
  \begin{block}{Proposition}
    For every \(x\in\Z\), one has \(x\le x\).
  \end{block}

  Choose \(0\in\N\) as the witness in the definition of \(\le\).  Since
  \(\iota(0)=0\),
  \[
    x=x+0=x+\iota(0).
  \]
  Therefore \(x\le x\).
\end{frame}

\begin{frame}{Transitivity of the integer order}
  Suppose \(x\le y\) and \(y\le z\).  Unpacking the two hypotheses gives
  \(p,q\in\N\) such that
  \[
    y=x+\iota(p),
    \qquad
    z=y+\iota(q).
  \]
  Substitute these equalities and use \(p+q\) as the new witness:
  \begin{align*}
    z
      &=(x+\iota(p))+\iota(q)\\
      &=x+\bigl(\iota(p)+\iota(q)\bigr)\\
      &=x+\iota(p+q).
  \end{align*}
  Hence \(x\le z\).
\end{frame}

\begin{frame}{Antisymmetry: the two gaps}
  Suppose \(x\le y\) and \(y\le x\).  Choose \(p,q\in\N\) such that
  \[
    y=x+\iota(p),
    \qquad
    x=y+\iota(q).
  \]
  Substitute \(y=x+\iota(p)\) in the second equality.  Then
  \[
    x=(x+\iota(p))+\iota(q)=x+\iota(p+q).
  \]
  Cancelling \(x\) from the additive equality yields
  \[
    \iota(p+q)=0=\iota(0).
  \]
\end{frame}

\begin{frame}{Antisymmetry: the gaps vanish}
  Since \(\iota\) is injective,
  \[
    \iota(p+q)=\iota(0)
    \quad\Longrightarrow\quad
    p+q=0.
  \]
  A sum of natural numbers is zero only when both summands are zero, so
  \[
    p=0,\qquad q=0.
  \]
  Therefore
  \[
    y=x+\iota(p)=x+\iota(0)=x.
  \]
  We have proved that \(x\le y\) and \(y\le x\) imply \(x=y\).
\end{frame}

\begin{frame}{Totality: the first natural comparison}
  \small
  Let
  \[
    x=\cl{a}{b},
    \qquad
    y=\cl{c}{d}.
  \]
  The natural numbers \(a+d\) and \(b+c\) are comparable:
  \[
    a+d\le b+c
    \qquad\text{or}\qquad
    b+c\le a+d.
  \]

  Suppose first that \(a+d\le b+c\).  By the definition of the order on
  \(\N\), choose \(e\in\N\) such that
  \[
    b+c=(a+d)+e
  \]
  and use \(e\) as a witness for \(x\le y\).  Indeed,
  \[
    x+\iota(e)=\cl{a+e}{b}.
  \]
  The displayed natural-number equality, after reassociation and
  commutation, is
  \[
    c+b=d+(a+e).
  \]
  Thus \(\cl{c}{d}=\cl{a+e}{b}=x+\iota(e)\), so \(x\le y\).
\end{frame}

\begin{frame}{Totality: the second natural comparison}
  \small
  Now suppose that \(b+c\le a+d\).  Choose \(e\in\N\) such that
  \[
    a+d=(b+c)+e.
  \]
  This time use \(e\) as a witness for \(y\le x\).  We have
  \[
    y+\iota(e)=\cl{c+e}{d},
  \]
  while reassociating the chosen equality gives
  \[
    a+d=b+(c+e).
  \]
  Hence
  \[
    \cl{a}{b}=\cl{c+e}{d}=y+\iota(e),
  \]
  and therefore \(y\le x\).  In either case, \(x\le y\) or \(y\le x\).
\end{frame}

\begin{frame}{The integers form a linear order}
  We have proved, directly from the witness definition of \(\le\), that
  for all \(x,y,z\in\Z\):
  \begin{align*}
    &x\le x,\\
    &x\le y\text{ and }y\le z \quad\Longrightarrow\quad x\le z,\\
    &x\le y\text{ and }y\le x \quad\Longrightarrow\quad x=y,\\
    &x\le y\text{ or }y\le x.
  \end{align*}

  \begin{block}{Conclusion}
    These are exactly the defining properties of a linear order.  Thus
    \((\Z,\le)\) is a linear order.
  \end{block}
\end{frame}

\begin{frame}{The embedding preserves and reflects order}
  \small
  \begin{block}{Proposition}
    For \(a,b\in\N\),
    \[
      \boxed{\iota(a)\le\iota(b)\Longleftrightarrow a\le b.}
    \]
  \end{block}

  Suppose \(\iota(a)\le\iota(b)\).  Choose \(n\in\N\) such that
  \[
    \iota(b)=\iota(a)+\iota(n)=\iota(a+n).
  \]
  Injectivity of \(\iota\) gives \(b=a+n\), hence \(a\le b\).

  Conversely, if \(a\le b\), choose \(k\in\N\) such that \(b=a+k\).  Then
  \[
    \iota(b)=\iota(a+k)=\iota(a)+\iota(k),
  \]
  so \(k\) witnesses \(\iota(a)\le\iota(b)\).
\end{frame}

\subsection{Order and arithmetic}

\begin{frame}{Addition preserves order}
  \begin{block}{Theorem}
    If \(x\le y\), then \(x+z\le y+z\).
  \end{block}

  Choose \(n\in\N\) such that \(y=x+\iota(n)\).  Then
  \begin{align*}
    y+z
      &=(x+\iota(n))+z\\
      &=(x+z)+\iota(n),
  \end{align*}
  where associativity and commutativity of addition justify the rearrangement.
  Thus the same natural gap \(n\) proves
  \[
    x+z\le y+z.
  \]
\end{frame}

\begin{frame}{Strict positivity}
  Since the order is linear, we write
  \[
    x<y
    \quad\Longleftrightarrow\quad
    x\le y\text{ and }x\ne y.
  \]

  If \(0<x\), then \(0\le x\), so the definition of order gives
  \[
    x=0+\iota(n)=\iota(n)
  \]
  for some \(n\in\N\).  Moreover \(n\ne0\), because \(x\ne0\) and \(\iota\)
  is injective.

  Conversely, if \(n\ne0\), then
  \[
    0<\iota(n).
  \]
  Indeed \(0\le\iota(n)\) has gap \(n\), and injectivity gives
  \(\iota(n)\ne0\).

  Therefore the positive integers are exactly the images of the positive
  natural numbers.
\end{frame}

\begin{frame}{A product of positive integers is positive}
  \begin{block}{Theorem}
    If \(0<x\) and \(0<y\), then \(0<xy\).
  \end{block}

  Write
  \[
    x=\iota(n),\qquad y=\iota(m),
    \qquad n,m\ne0.
  \]
  Since the embedding preserves multiplication,
  \[
    xy=\iota(n)\iota(m)=\iota(nm).
  \]
  The natural-number product \(nm\) is nonzero.  The characterization of
  positive integers therefore gives
  \[
    0<\iota(nm)=xy.
  \]
\end{frame}

\begin{frame}{An ordered commutative ring}
  We have shown that
  \begin{itemize}
    \item \(\Z\) is a commutative ring;
    \item \(\le\) is a linear order;
    \item adding the same integer preserves order;
    \item the product of two positive integers is positive.
  \end{itemize}

  These facts make \(\Z\) an \alert{ordered commutative ring}.

  They also recover all the familiar sign behavior.  For example, translating
  \(x<y\) by \(-x\) gives
  \[
    0<y-x.
  \]
\end{frame}

\begin{frame}{The Archimedean bound}
  \begin{block}{Theorem}
    For every \(x\in\Z\), there is an \(n\in\N\) such that
    \[
      x\le\iota(n).
    \]
  \end{block}

  Let \(x=\cl{a}{b}\).  We claim that \(n=a\) works.  Indeed,
  \[
    x+\iota(b)
      =\cl{a}{b}+\cl{b}{0}
      =\cl{a+b}{b}.
  \]
  But
  \[
    \cl{a+b}{b}=\cl{a}{0}=\iota(a),
  \]
  because \((a+b)+0=b+a\).  Thus
  \[
    \iota(a)=x+\iota(b),
  \]
  and the natural gap \(b\) proves \(x\le\iota(a)\).
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  Starting from \(\N\), we constructed \(\Z\) and proved that
  \begin{itemize}
    \item integers are equivalence classes of pre-integers, that is, of
      formal differences;
    \item addition, negation, and multiplication are independent of
      representatives;
    \item \(\Z\) is a commutative ring with no zero divisors;
    \item nonzero factors may be cancelled;
    \item \(\N\) embeds into \(\Z\), preserving arithmetic and order;
    \item the nonnegative-gap relation is a linear order;
    \item addition preserves order and positive products are positive;
    \item every integer is bounded above by a natural number.
  \end{itemize}
\end{frame}


\begin{frame}{An equation without solutions in \(\Z\)}
  Suppose that some \(x\in\Z\) satisfied
  \[
    2x=1.
  \]
  By totality of the integer order,
  \[
    0\le x
    \qquad\text{or}\qquad
    x\le0.
  \]

  If \(0\le x\), write \(x=\iota(n)\) with \(n\in\N\).  Since \(\iota\)
  preserves multiplication and is injective,
  \[
    \iota(2n)=2\iota(n)=2x=1=\iota(1).
  \]
  Hence
  \[
    2n=1.
  \]

  If \(x\le0\), write \(0=x+\iota(n)\) with \(n\in\N\).  Multiplying by
  \(2\) and again using injectivity gives
  \[
    0=2x+\iota(2n)=1+\iota(2n)=\iota(1+2n),
  \]
  and hence
  \[
    0=1+2n.
  \]
\end{frame}

\begin{frame}{An equation without solutions in \(\Z\) (continued)}
  We now rule out the two natural-number equations exactly as before, using
  the recursive description of addition and the successor principles.

  First,
  \[
    1+2n=2n+1=\suc{(2n)},
  \]
  so \(0=1+2n\) is impossible because zero is not a successor.

  Now suppose that \(2n=1\).  If \(n=0\), this says \(0=\suc0\), which is
  impossible.  If \(n=\suc m\), then
  \[
    2n=2(m+1)=2m+2=\suc{\suc{(2m)}}.
  \]
  Hence \(2n=1=\suc0\) would imply, by injectivity of the successor
  operation,
  \[
    \suc{(2m)}=0,
  \]
  again impossible because zero is not a successor.

  Therefore neither natural-number equation can hold, and \(2x=1\) has no
  solution in \(\Z\).
\end{frame}

\section{The rational numbers}

\subsection{Fractions as equivalence classes}

\begin{frame}{Why introduce the rational numbers?}
  In \(\Z\), the equation
  \[
    2x=1
  \]
  has no solution.  To solve it we want to introduce fractions \(a/b\), where
  \(a,b\in\Z\) and \(b\ne0\).

  We begin with a pair \((a,b)\), intended to represent \(a/b\).

  \begin{alertblock}{The central difficulty}
    A rational number has many representatives:
    \[
      \frac12=\frac24=\frac{-3}{-6}.
    \]
    Before doing arithmetic, we must say exactly when two pairs represent the
    same number.
  \end{alertblock}
\end{frame}

\begin{frame}{The pre-rationals}
  \begin{block}{Definition}
    The set of \alert{pre-rationals} is
    \[
      \Q_{\mathrm{pre}}
        =\{(a,b)\in\Z^2:b\ne0\}.
    \]
  \end{block}

  The restriction on the denominator is essential.  It ensures that products
  of denominators remain nonzero and, later, that cross multiplication admits
  cancellation.

  We shall use fraction notation for the pair when no confusion can arise:
  \[
    (a,b)\quad\longleftrightarrow\quad \frac ab.
  \]
\end{frame}

\begin{frame}{When do two fractions agree?}
  \begin{block}{Definition}
    For \((a,b),(c,d)\in\Q_{\mathrm{pre}}\), declare
    \[
      (a,b)\sim(c,d)
      \quad\Longleftrightarrow\quad
      ad=bc.
    \]
  \end{block}

  This uses only multiplication and equality in \(\Z\), but it captures the
  familiar rule
  \[
    \frac ab=\frac cd
    \quad\Longleftrightarrow\quad ad=bc.
  \]

  For example, \((1,2)\sim(-3,-6)\) because
  \(1\cdot(-6)=2\cdot(-3)\).
\end{frame}

\begin{frame}{Reflexivity and symmetry}
  \begin{block}{Proposition}
    The relation \(\sim\) is reflexive and symmetric.
  \end{block}

  \begin{proof}
    For every \((a,b)\), commutativity gives \(ab=ba\), hence
    \((a,b)\sim(a,b)\).

    If \((a,b)\sim(c,d)\), then \(ad=bc\).  Reading the equality backwards
    and commuting both products gives
    \[
      cb=da,
    \]
    which says \((c,d)\sim(a,b)\).
  \end{proof}
\end{frame}

\begin{frame}{Transitivity uses the nonzero denominator}
  \small
  Suppose
  \[
    (a,b)\sim(c,d),
    \qquad
    (c,d)\sim(e,f).
  \]
  Thus \(ad=bc\) and \(cf=de\).  Multiplying the first equality by \(f\)
  and the second by \(b\) gives
  \[
    adf=bcf=bde.
  \]
  After rearranging,
  \[
    d(af)=d(be).
  \]
  Since \(d\ne0\), cancellation in \(\Z\) yields \(af=be\).  Therefore
  \((a,b)\sim(e,f)\).

  \begin{alertblock}{Why \(b\ne0\) belongs in the definition}
    Without a nonzero middle denominator, this cancellation argument would
    fail and transitivity would not follow.
  \end{alertblock}
\end{frame}

\begin{frame}{The rational numbers}
  We have proved that \(\sim\) is an equivalence relation.

  \begin{block}{Definition}
    The rational numbers are the equivalence classes
    \[
      \boxed{\Q=\Q_{\mathrm{pre}}/{\sim}}.
    \]
    We write
    \[
      \qcl ab
    \]
    for the class represented by \((a,b)\).
  \end{block}

  Equality in \(\Q\) is therefore governed by the single criterion
  \[
    \boxed{\qcl ab=\qcl cd\quad\Longleftrightarrow\quad ad=bc.}
  \]
\end{frame}

\subsection{Arithmetic on rational numbers}

\begin{frame}{The four basic operations}
  The usual fraction formulas suggest
  \begin{align*}
    -\qcl ab&=\qcl{-a}b,\\
    \qcl ab+\qcl cd&=\qcl{ad+bc}{bd},\\
    \qcl ab\qcl cd&=\qcl{ac}{bd}.
  \end{align*}
  If \(\qcl ab\ne0\), equivalently \(a\ne0\), we define its reciprocal by
  \[
    \left(\qcl ab\right)^{-1}
      =\qcl ba.
  \]

  All displayed denominators are nonzero: this follows from the absence of
  zero divisors in \(\Z\), and from \(a\ne0\) in the reciprocal formula.  The
  next task is to prove that the answers do not depend on the chosen
  representatives.
\end{frame}

\begin{frame}{Addition is well defined}
  \footnotesize
  Suppose
  \[
    \frac ab\sim\frac{a'}{b'},
    \qquad
    \frac cd\sim\frac{c'}{d'},
  \]
  so \(ab'=ba'\) and \(cd'=dc'\).  We compare the proposed sums by cross
  multiplication:
  \begin{align*}
    (ad+bc)b'd'
      &=ab'dd'+bb'cd'\\
      &=ba'dd'+bb'dc'\\
      &=bd(a'd'+b'c').
  \end{align*}
  Hence
  \[
    \frac{ad+bc}{bd}\sim
    \frac{a'd'+b'c'}{b'd'}.
  \]

  Thus addition passes from pre-rationals to equivalence classes.
\end{frame}

\begin{frame}{Multiplication is well defined}
  \small
  Under the same hypotheses \(ab'=ba'\) and \(cd'=dc'\),
  \begin{align*}
    (ac)(b'd')
      &=(ab')(cd')\\
      &=(ba')(dc')\\
      &=(bd)(a'c').
  \end{align*}
  Consequently,
  \[
    \frac{ac}{bd}\sim\frac{a'c'}{b'd'}.
  \]

  Negation is even more direct: from \(ad=bc\) we obtain
  \[
    (-a)d=b(-c),
  \]
  so \((-a,b)\sim(-c,d)\).

  Therefore multiplication and negation are well defined on \(\Q\).
\end{frame}

\begin{frame}{The zero class detects a zero numerator}
  \begin{block}{Proposition}
    If \(b\ne0\), then
    \[
      \qcl ab=\qcl01
      \quad\Longleftrightarrow\quad a=0.
    \]
  \end{block}

  \begin{proof}
    The equality criterion says
    \[
      a\cdot1=b\cdot0,
    \]
    which is equivalent to \(a=0\).
  \end{proof}

  In particular, a nonzero rational has a nonzero numerator in every
  representative.  This is exactly what is needed to define its reciprocal.
\end{frame}

\begin{frame}{The reciprocal is well defined}
  Suppose \((a,b)\sim(c,d)\) and that their common class is nonzero.  Then
  \(ad=bc\), and the preceding proposition gives
  \[
    a\ne0
    \qquad\text{and}\qquad
    c\ne0.
  \]
  Thus \((b,a)\) and \((d,c)\) are valid pre-rationals.  Moreover,
  \[
    (b,a)\sim(d,c)
    \quad\Longleftrightarrow\quad
    bc=ad,
  \]
  which is our original equality in reverse.

  Therefore \(x^{-1}=\qcl ba\) is independent of the representative chosen
  for every nonzero \(x\in\Q\).
\end{frame}

\begin{frame}{Zero, one, and subtraction}
  We define
  \[
    0=\qcl01,
    \qquad
    1=\qcl11.
  \]
  Subtraction is addition of the negative.  Directly from the fraction
  formulas,
  \[
    \boxed{
      \qcl ab-\qcl cd=\qcl{ad-bc}{bd}.}
  \]

  Also \(0\ne1\): otherwise the equality criterion would give
  \[
    0\cdot1=1\cdot1,
  \]
  contradicting \(0\ne1\) in \(\Z\).
\end{frame}

\begin{frame}{The commutative-ring laws}
  \small
  Every ring identity can now be checked on representatives.  For example,
  \begin{align*}
    \left(\qcl ab+\qcl cd\right)+\qcl ef
      &=\qcl{(ad+bc)f+(bd)e}{bdf}\\
      &=\qcl{adf+bcf+bde}{bdf},\\[0.3em]
    \qcl ab+\left(\qcl cd+\qcl ef\right)
      &=\qcl{a(df)+b(cf+de)}{bdf}\\
      &=\qcl{adf+bcf+bde}{bdf}.
  \end{align*}

  Associativity and commutativity of both operations, the identity laws,
  additive inverses, and distributivity all reduce in this way to polynomial
  identities in \(\Z\).

  \begin{block}{Conclusion}
    The rational numbers form a nontrivial commutative ring.
  \end{block}
\end{frame}

\begin{frame}{Multiplying by the reciprocal}
  \begin{block}{Proposition}
    If \(x\in\Q\) and \(x\ne0\), then \(xx^{-1}=1\).
  \end{block}

  \begin{proof}
    Write \(x=\qcl ab\).  Since \(x\ne0\), its numerator satisfies
    \(a\ne0\).  Therefore
    \[
      xx^{-1}
        =\qcl ab\qcl ba
        =\qcl{ab}{ba}
        =\qcl11,
    \]
    where the last equality follows from
    \((ab)\cdot1=(ba)\cdot1\).
  \end{proof}
\end{frame}

\begin{frame}{The field of rational numbers}
  We have constructed:
  \begin{itemize}
    \item a nontrivial commutative ring \(\Q\);
    \item a reciprocal \(x^{-1}\) for every nonzero rational \(x\);
    \item the identity \(xx^{-1}=1\) for every nonzero rational \(x\).
  \end{itemize}

  \begin{block}{Theorem}
    With these operations, \(\Q\) is a field.
  \end{block}

  Thus for \(y\ne0\), division can be defined by
  \[
    \frac{x}{y}=xy^{-1},
  \]
  and equations such as \(2x=1\) now have the expected solution
  \(x=2^{-1}\).
\end{frame}

\subsection{The integers inside the rationals}

\begin{frame}{The integer embedding}
  \begin{block}{Definition}
    Define
    \[
      j\colon\Z\longrightarrow\Q,
      \qquad
      j(a)=\qcl a1.
    \]
  \end{block}

  The fraction formulas show that
  \begin{align*}
    j(0)&=0,& j(1)&=1,\\
    j(a+c)&=j(a)+j(c),&
    j(ac)&=j(a)j(c).
  \end{align*}

  Hence the old integer arithmetic is preserved inside the new number system.
\end{frame}

\begin{frame}{No integers are identified}
  \begin{block}{Proposition}
    The map \(j\colon\Z\to\Q\) is injective.
  \end{block}

  \begin{proof}
    If \(j(a)=j(c)\), then
    \[
      \qcl a1=\qcl c1.
    \]
    Cross multiplication gives \(a\cdot1=1\cdot c\), hence \(a=c\).
  \end{proof}

  We may therefore regard \(\Z\) as a subring of \(\Q\), although the
  construction keeps the embedding visible.
\end{frame}

\begin{frame}{Every fraction is a quotient of integers}
  \begin{block}{Proposition}
    For \(a,b\in\Z\) with \(b\ne0\),
    \[
      \boxed{\qcl ab=j(a)j(b)^{-1}.}
    \]
  \end{block}

  Indeed,
  \[
    j(a)j(b)^{-1}
      =\qcl a1\qcl1b
      =\qcl ab.
  \]

  This formula says that the quotient construction has done exactly what was
  intended: it adjoins inverses of all nonzero integers, and every rational is
  obtained from those inverses.
\end{frame}

\begin{frame}{The natural embedding is compatible}
  Let \(\iota\colon\N\to\Z\) be the embedding already constructed for the integers.
  Define
  \[
    i\colon\N\longrightarrow\Q,
    \qquad
    i(n)=\qcl{\iota(n)}1.
  \]
  Then
  \[
    i=j\circ\iota.
  \]

  Since both earlier embeddings preserve \(0,1,+,\cdot\) and are injective,
  the same is true of \(i\).  Thus the chain
  \[
    \N\lhook\joinrel\longrightarrow
    \Z\lhook\joinrel\longrightarrow
    \Q
  \]
  preserves the arithmetic constructed at every previous stage.
\end{frame}

\subsection{Nonnegativity}

\begin{frame}{Choosing a positive denominator}
  \begin{block}{Proposition}
    Every rational number has a representative \(\qcl ab\) with \(b>0\).
  \end{block}

  \begin{proof}
    Begin with any representative \(\qcl ab\), where \(b\ne0\).  The total
    order on \(\Z\) gives either \(b>0\) or \(b<0\).

    In the first case there is nothing to do.  In the second case, \(-b>0\)
    and
    \[
      \qcl ab=\qcl{-a}{-b},
    \]
    because \(a(-b)=b(-a)\).
  \end{proof}

  Positive denominators let the sign of a rational be read from its numerator.
\end{frame}

\begin{frame}{Definition of nonnegativity}
  \begin{block}{Definition}
    A rational \(x\) is \alert{nonnegative} if there are integers \(a,b\) such
    that
    \[
      a\ge0,\qquad b>0,
      \qquad x=\qcl ab.
    \]
  \end{block}

  We write this property temporarily as \(x\in\Q_{\ge0}\).  It is independent
  of any single chosen representative: it asks only for the existence of one
  representative whose numerator and denominator have the required signs.

  Clearly \(0,1\in\Q_{\ge0}\), represented by \(0/1\) and \(1/1\).
\end{frame}

\begin{frame}{Every rational has one of the two signs}
  \begin{block}{Proposition}
    For every \(x\in\Q\), either \(x\in\Q_{\ge0}\) or
    \(-x\in\Q_{\ge0}\).
  \end{block}

  \begin{proof}
    Choose \(x=\qcl ab\) with \(b>0\).  By totality in \(\Z\), either
    \(a\ge0\) or \(a<0\).

    In the first case, \(a/b\) witnesses that \(x\) is nonnegative.  In the
    second, \(-a>0\), and
    \[
      -x=\qcl{-a}b
    \]
    is nonnegative.
  \end{proof}
\end{frame}

\begin{frame}{Opposite signs meet only at zero}
  \small
  \begin{block}{Proposition}
    If both \(x\) and \(-x\) are nonnegative, then \(x=0\).
  \end{block}

  \begin{proof}
    Choose
    \[
      x=\qcl ab,
      \qquad
      -x=\qcl cd,
    \]
    with \(a,c\ge0\) and \(b,d>0\).  Since
    \(-x=\qcl{-a}b\), equality of the second pair of representatives gives
    \[
      (-a)d=bc.
    \]
    The left side is nonpositive and the right side is nonnegative, so both
    are zero.  Since \(d>0\), this forces \(a=0\), and hence \(x=0\).
  \end{proof}
\end{frame}

\begin{frame}{Nonnegativity is closed under addition}
  Suppose
  \[
    x=\qcl ab,
    \qquad
    y=\qcl cd,
  \]
  where \(a,c\ge0\) and \(b,d>0\).  Then
  \[
    x+y=\qcl{ad+bc}{bd}.
  \]

  Products of nonnegative integers are nonnegative, so
  \(ad+bc\ge0\).  Products of positive integers are positive, so \(bd>0\).

  \begin{block}{Conclusion}
    If \(x,y\in\Q_{\ge0}\), then \(x+y\in\Q_{\ge0}\).
  \end{block}
\end{frame}

\begin{frame}{Nonnegativity and multiplication}
  With the same positive-denominator representatives,
  \[
    xy=\qcl{ac}{bd}.
  \]
  Since \(ac\ge0\) and \(bd>0\), the product is nonnegative.

  If \(x\ne0\), then \(a>0\), and
  \[
    x^{-1}=\qcl ba
  \]
  is positive because \(b>0\) and \(a>0\).

  \begin{block}{Conclusion}
    Products of nonnegative rationals are nonnegative, and reciprocals of
    positive rationals are positive.
  \end{block}
\end{frame}

\subsection{The order relation}

\begin{frame}{Order as a nonnegative gap}
  The order on \(\Z\) was defined by asking whether one integer differs from
  another by a nonnegative amount.  We use the same idea here.

  \begin{block}{Definition}
    For \(x,y\in\Q\), define
    \[
      \boxed{x\le y
        \quad\Longleftrightarrow\quad
        y-x\in\Q_{\ge0}.}
    \]
  \end{block}

  In particular,
  \[
    0\le x
    \quad\Longleftrightarrow\quad
    x\in\Q_{\ge0},
  \]
  because \(x-0=x\).
\end{frame}

\begin{frame}{Comparing positive-denominator fractions}
  \small
  \begin{block}{Proposition}
    If \(b,d>0\), then
    \[
      \boxed{
      \qcl ab\le\qcl cd
      \quad\Longleftrightarrow\quad
      ad\le bc.}
    \]
  \end{block}

  Indeed,
  \[
    \qcl cd-\qcl ab=\qcl{bc-ad}{bd},
  \]
  and \(bd>0\).  Thus the gap is nonnegative exactly when
  \(bc-ad\ge0\), which is equivalent to \(ad\le bc\).

  This recovers the ordinary cross-multiplication rule without ever reversing
  an inequality, because the denominators have first been made positive.
\end{frame}

\begin{frame}{Reflexivity and transitivity}
  \begin{block}{Proposition}
    The relation \(\le\) on \(\Q\) is reflexive and transitive.
  \end{block}

  For reflexivity,
  \[
    x-x=0\in\Q_{\ge0},
  \]
  so \(x\le x\).

  If \(x\le y\) and \(y\le z\), then \(y-x\) and \(z-y\) are
  nonnegative.  Their sum is nonnegative, and
  \[
    (y-x)+(z-y)=z-x.
  \]
  Therefore \(x\le z\).
\end{frame}

\begin{frame}{Antisymmetry}
  \begin{block}{Proposition}
    If \(x\le y\) and \(y\le x\), then \(x=y\).
  \end{block}

  \begin{proof}
    The two assumptions say that
    \[
      y-x\in\Q_{\ge0},
      \qquad
      x-y\in\Q_{\ge0}.
    \]
    But \(x-y=-(y-x)\).  A rational and its negative can both be
    nonnegative only when that rational is zero.  Hence \(y-x=0\), and
    therefore \(x=y\).
  \end{proof}
\end{frame}

\begin{frame}{Totality}
  \begin{block}{Proposition}
    For all \(x,y\in\Q\), either \(x\le y\) or \(y\le x\).
  \end{block}

  \begin{proof}
    Apply the sign dichotomy to the gap \(y-x\).  Either
    \[
      y-x\in\Q_{\ge0},
    \]
    which means \(x\le y\), or
    \[
      -(y-x)=x-y\in\Q_{\ge0},
    \]
    which means \(y\le x\).
  \end{proof}

  Together with reflexivity, transitivity, and antisymmetry, this makes
  \(\Q\) a linearly ordered set.
\end{frame}

\subsection{Order and field arithmetic}

\begin{frame}{Translation preserves order}
  \begin{block}{Proposition}
    If \(x\le y\), then \(x+t\le y+t\) for every \(t\in\Q\).
  \end{block}

  \begin{proof}
    The relevant gap does not change:
    \[
      (y+t)-(x+t)=y-x.
    \]
    Since the right side is nonnegative, so is the left side.
  \end{proof}

  Consequently, adding the same rational to both sides preserves and reflects
  both weak and strict inequalities.
\end{frame}

\begin{frame}{Products of nonnegative rationals}
  \begin{block}{Proposition}
    If \(0\le x\) and \(0\le y\), then \(0\le xy\).
  \end{block}

  This is precisely closure of \(\Q_{\ge0}\) under multiplication, translated
  through the equivalence
  \[
    0\le x\quad\Longleftrightarrow\quad x\in\Q_{\ge0}.
  \]

  Combining this fact with translation invariance yields the familiar rule:
  if \(x\le y\) and \(0\le t\), then
  \[
    xt\le yt.
  \]
\end{frame}

\begin{frame}{The embeddings preserve and reflect order}
  \begin{block}{Proposition}
    For \(p,q\in\Z\),
    \[
      j(p)\le j(q)
      \quad\Longleftrightarrow\quad
      p\le q.
    \]
  \end{block}

  Both fractions have denominator \(1>0\), so the comparison criterion gives
  \[
    \qcl p1\le\qcl q1
      \quad\Longleftrightarrow\quad
      p\cdot1\le1\cdot q.
  \]

  Since \(i=j\circ\iota\), the embedding \(i\colon\N\to\Q\) likewise preserves
  and reflects \(\le\) and \(<\).
\end{frame}

\begin{frame}{Strictly positive products}
  \begin{block}{Proposition}
    If \(0<x\) and \(0<y\), then \(0<xy\).
  \end{block}

  Choose positive-denominator representatives
  \[
    x=\qcl ab,
    \qquad
    y=\qcl cd.
  \]
  Strict positivity means \(a>0\) and \(c>0\).  Hence \(ac>0\), while
  \(bd>0\), so
  \[
    xy=\qcl{ac}{bd}>0.
  \]

  Thus multiplication by a positive rational strictly preserves order.
\end{frame}

\begin{frame}{An ordered field}
  We have now proved that \(\Q\) is simultaneously
  \begin{itemize}
    \item a field;
    \item a linear order;
    \item ordered compatibly with addition;
    \item ordered compatibly with multiplication by nonnegative elements.
  \end{itemize}

  \begin{block}{Theorem}
    The constructed rational numbers form a linearly ordered field.
  \end{block}

  Algebra and order are not separate structures here: both ultimately descend
  from the arithmetic and order of \(\Z\).
\end{frame}

\begin{frame}{The Archimedean bound: reducing to integers}
  \small
  \begin{block}{Theorem}
    For every \(x\in\Q\), there exists \(n\in\N\) such that \(x\le i(n)\).
  \end{block}

  If \(x<0\), choose \(n=0\).  Otherwise choose a representation
  \[
    x=\qcl ab
  \]
  with \(a\ge0\) and \(b>0\).

  The Archimedean property of \(\Z\) supplies \(n\in\N\) with
  \[
    a\le\iota(n).
  \]
  Since \(b\) is a positive integer, \(1\le b\), and since
  \(\iota(n)\ge0\),
  \[
    \iota(n)\le b\,\iota(n).
  \]
\end{frame}

\begin{frame}{The Archimedean bound: finishing the comparison}
  Combining the two integer inequalities gives
  \[
    a\le b\,\iota(n).
  \]
  Both denominators below are positive, so the cross-multiplication criterion
  yields
  \[
    \qcl ab
      \le
    \qcl{\iota(n)}1=i(n).
  \]

  \begin{alertblock}{What the theorem rules out}
    No rational number is larger than every natural number.  The field \(\Q\)
    contains arbitrarily large numbers.
  \end{alertblock}
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  We proved that
  \begin{itemize}
    \item rational numbers are equivalence classes of fractions, under the
      cross-multiplication criterion;
    \item \(\Q\) is a field, and every element has the form
      \(j(a)j(b)^{-1}\) with \(a,b\in\Z\) and \(b\ne0\);
    \item \(\Q\) is linearly ordered; addition preserves order, and products
      of nonnegative elements are nonnegative;
    \item the embeddings \(\N\hookrightarrow\Z\hookrightarrow\Q\) preserve
      and reflect arithmetic and order;
    \item \(\Q\) satisfies the Archimedean bound.
  \end{itemize}
\end{frame}

\section{\texorpdfstring{Arithmetic in \(\N\)}{Arithmetic in N}}

\begin{frame}{Why study arithmetic in \(\N\)?}
  The next construction, the real numbers, is motivated by an equation that
  cannot be solved in \(\Q\):
  \[
    x^2=2.
  \]
  Proving this impossibility requires genuinely arithmetic tools, which this
  section develops:
  \begin{enumerate}
    \item \alert{Euclidean division}, with a unique quotient and remainder;
    \item the \alert{greatest common divisor} and Bézout's identity;
    \item \alert{Euclid's lemma}: a prime dividing a product divides a factor;
    \item the \alert{fundamental theorem of arithmetic}: every number from
      \(2\) on is a product of primes in exactly one way.
  \end{enumerate}
  Counting the prime factors on both sides of \(A^2=2B^2\) will then produce
  the desired contradiction.
\end{frame}

\subsection{Euclidean division}

\begin{frame}{Euclidean division}
  \begin{block}{Theorem}
    Let \(n,d\in\N\) with \(d>0\).  There are unique \(q,r\in\N\) such that
    \[
      \boxed{n=dq+r,\qquad r<d.}
    \]
  \end{block}

  The number \(q\) is the quotient and \(r\) is the remainder.  The bound
  \(r<d\) is what singles them out among all decompositions of \(n\).

  \begin{alertblock}{Proof strategy}
    Keep the positive divisor \(d\) fixed and use induction on the dividend
    \(n\).  Passing from \(n\) to \(\suc n\) either increments the remainder
    or resets it to zero and increments the quotient.
  \end{alertblock}
\end{frame}

\begin{frame}{Euclidean division: the induction}
  For \(n=0\), take \(q=r=0\).  Then
  \[
    0=d\cdot0+0,
    \qquad
    0<d.
  \]

  Now suppose
  \[
    n=dq+r,
    \qquad
    r<d.
  \]
  Since \(r<d\), the gap from \(r\) to \(d\) is positive: \(d=r+\suc y\)
  for some \(y\in\N\).  Moving the successor through the sum gives
  \(d=\suc r+y\), so \(\suc r\le d\).  Hence there are two possibilities:
  \[
    \suc r<d
    \qquad\text{or}\qquad
    \suc r=d.
  \]

  These alternatives determine the quotient and remainder for \(\suc n\).
\end{frame}

\begin{frame}{Euclidean division: the two cases}
  If \(\suc r<d\), then
  \[
    \suc n
      =\suc{(dq+r)}
      =dq+\suc r.
  \]
  Thus the quotient stays \(q\) and the new remainder is \(\suc r\).

  If \(\suc r=d\), then
  \[
    \suc n
      =dq+\suc r
      =dq+d
      =d(q+1)+0.
  \]
  Thus the quotient becomes \(q+1\) and the new remainder is \(0<d\).

  In either case \(\suc n\) has the required decomposition, so induction
  proves existence for every \(n\in\N\).
\end{frame}

\begin{frame}{Uniqueness of quotient and remainder}
  Suppose
  \[
    n=dq+r=dq'+r',
    \qquad
    r,r'<d.
  \]
  If \(q<q'\), write \(q'=q+s\) with \(s>0\).  Writing \(s=\suc t\) gives
  \(ds=dt+d\ge d\), while distributivity gives \(dq'=dq+ds\).  Consequently,
  \[
    n=dq+r<dq+d\le dq+ds=dq'\le dq'+r'=n,
  \]
  a contradiction.  The same argument rules out \(q'<q\), so totality gives
  \(q=q'\).

  Addition cancellation in \(dq+r=dq+r'\) now gives \(r=r'\).  This proves
  uniqueness.
\end{frame}

\subsection{Prime numbers and Euclid's lemma}

\begin{frame}{Divisibility and prime numbers}
  For \(a,b\in\N\), define
  \[
    a\mid b
    \quad\Longleftrightarrow\quad
    b=ac\ \text{ for some }c\in\N.
  \]

  \begin{block}{Definition}
    A number \(p\in\N\) is \alert{prime} if \(2\le p\) and
    \[
      d\mid p
      \quad\Longrightarrow\quad
      d=1\ \text{ or }\ d=p.
    \]
  \end{block}

  Thus \(0\) and \(1\) are not prime, while \(2,3,5,7\) are prime.  A number
  \(n\ge2\) that is not prime is \alert{composite}; it can be written
  \[
    n=ab
    \qquad\text{with}\qquad
    2\le a<n,\qquad 2\le b<n.
  \]
\end{frame}

\begin{frame}{The greatest common divisor}
  Let \(a,b\in\N\), not both zero.

  \begin{block}{Definition}
    The \alert{greatest common divisor} of \(a\) and \(b\) is the natural
    number \(g=\gcd(a,b)\) characterized by
    \begin{enumerate}
      \item \(g\mid a\) and \(g\mid b\);
      \item if \(d\mid a\) and \(d\mid b\), then \(d\mid g\).
    \end{enumerate}
  \end{block}

  Thus \(g\) is a common divisor that is divisible by every common divisor.
  Two numbers with these properties divide each other and are therefore
  equal, so the two conditions determine \(g\) uniquely; the Euclidean
  algorithm below produces one.  Since divisibility respects the natural
  order for positive numbers, \(g\) is also the largest common divisor in
  the usual order.

  For example,
  \[
    \gcd(84,30)=6.
  \]
\end{frame}

\begin{frame}{The Euclidean algorithm}
  Euclidean division can be repeated.  For \(a,b>0\), begin with
  \begin{align*}
    a&=bq_0+r_0,       &0\le r_0&<b,\\
    b&=r_0q_1+r_1,     &0\le r_1&<r_0,\\
    r_0&=r_1q_2+r_2,   &0\le r_2&<r_1,
  \end{align*}
  and continue while the remainder is nonzero.

  The remainders strictly decrease, and a strictly decreasing sequence of
  natural numbers must terminate.  Each step preserves common divisors:
  since \(a=bq_0+r_0\), every common divisor of \(b\) and \(r_0\) divides
  \(a\), and every common divisor of \(a\) and \(b\) divides \(r_0=a-bq_0\).

  \begin{block}{Conclusion}
    The last nonzero remainder is \(\gcd(a,b)\), the greatest common divisor
    of \(a\) and \(b\).
  \end{block}
\end{frame}

\begin{frame}{Bézout's identity}
  Each line of the Euclidean algorithm can be rearranged in \(\Z\):
  \[
    r_{i+1}=r_{i-1}-q_{i+1}r_i.
  \]
  Starting with the last nonzero remainder and substituting backwards
  expresses it as an integer linear combination of the original inputs.

  \begin{block}{Bézout's identity}
    For \(a,b>0\), there are \(u,v\in\Z\) such that
    \[
      \boxed{\gcd(a,b)=ua+vb.}
    \]
    Here the natural numbers are viewed inside \(\Z\).
  \end{block}

  In particular, if \(\gcd(a,b)=1\), then
  \[
    ua+vb=1
  \]
  for suitable integers \(u,v\).
\end{frame}

\begin{frame}{Euclid's lemma}
  \begin{block}{Lemma}
    If \(p\) is prime and \(p\mid ab\), then
    \[
      \boxed{p\mid a\quad\text{or}\quad p\mid b.}
    \]
  \end{block}

  Suppose \(p\nmid a\).  Every common divisor of \(p\) and \(a\) divides
  the prime \(p\), so it is \(1\) or \(p\).  The second possibility is
  excluded, hence \(\gcd(p,a)=1\).

  Bézout's identity gives \(up+va=1\).  Multiplying by \(b\),
  \[
    ubp+vab=b.
  \]
  Both terms on the left are divisible by \(p\): the first visibly, and the
  second because \(p\mid ab\).  Therefore \(p\mid b\), proving the lemma.
\end{frame}

\subsection{The fundamental theorem of arithmetic}

\begin{frame}{The fundamental theorem of arithmetic}
  \begin{block}{Theorem}
    Every \(n\in\N\) with \(n\ge2\) is a product of primes:
    \[
      n=p_1p_2\cdots p_k.
    \]
    This factorization is unique up to reordering the prime factors.
  \end{block}

  Equivalently, if
  \[
    n=p_1\cdots p_k=q_1\cdots q_\ell
  \]
  and both lists are arranged in nondecreasing order, then
  \[
    k=\ell
    \qquad\text{and}\qquad
    p_i=q_i\quad\text{for every }i.
  \]

  Existence comes from induction; uniqueness comes from Euclid's lemma.
\end{frame}

\begin{frame}{Existence of prime factorization}
  We use strong induction on \(n\), which follows from the induction principle
  for \(\N\).

  If \(n\) is prime, it is already a product consisting of one prime.  If it
  is not prime, then
  \[
    n=ab
    \qquad\text{with}\qquad
    2\le a<n,\qquad 2\le b<n.
  \]
  The induction hypothesis supplies prime factorizations
  \[
    a=p_1\cdots p_k,
    \qquad
    b=q_1\cdots q_\ell.
  \]
  Multiplying them gives
  \[
    n=p_1\cdots p_kq_1\cdots q_\ell.
  \]

  Thus every \(n\ge2\) has a prime factorization.
\end{frame}

\begin{frame}{Uniqueness of prime factorization}
  Suppose
  \[
    p_1\cdots p_k=q_1\cdots q_\ell.
  \]
  The prime \(p_1\) divides the product on the right.  Repeated application of
  Euclid's lemma shows that \(p_1\mid q_j\) for some \(j\).  Since \(q_j\) is
  prime and \(p_1\ne1\), this forces
  \[
    p_1=q_j.
  \]

  Reorder the right-hand factors so that \(q_j\) comes first and cancel this
  common nonzero factor.  Repeating the argument matches and cancels every
  prime on both sides.

  The process cannot exhaust one side first: once every factor on one side
  is cancelled, the other side is a product of primes equal to \(1\), which
  is impossible because every prime is at least \(2\).

  Therefore the two lists contain exactly the same primes with exactly the
  same multiplicities.  This proves uniqueness up to reordering.
\end{frame}

\subsection{An application to the rational numbers}

\begin{frame}{Prime multiplicities in a square}
  For a prime \(p\) and a positive natural number \(n\), let
  \(\nu_p(n)\) be the number of times \(p\) occurs in the unique prime
  factorization of \(n\); the factorization of \(1\) is empty, so
  \(\nu_p(1)=0\).

  Concatenating factorizations of \(m\) and \(n\) yields a factorization of
  \(mn\); by uniqueness it is \emph{the} factorization, so multiplicities
  add:
  \[
    \nu_p(mn)=\nu_p(m)+\nu_p(n).
  \]
  In particular,
  \[
    \boxed{\nu_p(n^2)=2\nu_p(n).}
  \]

  Thus every prime occurs an even number of times in the factorization of a
  square.  For the prime \(2\), this says that \(\nu_2(n^2)\) is even.
\end{frame}

\begin{frame}{The equation \(x^2=2\) in \(\Q\)}
  Suppose that \(x\in\Q\) satisfied \(x^2=2\).  Replacing \(x\) by \(-x\)
  if necessary, we may assume \(x\ge0\).  A positive-denominator
  representative then has the form
  \[
    x=\qcl{\iota(A)}{\iota(B)},
    \qquad
    A,B\in\N,\qquad B>0.
  \]

  Squaring and using the equality criterion for rational numbers gives
  \[
    A^2=2B^2
  \]
  in \(\N\).  Since \(B>0\), the right side is positive; hence \(A\ne0\)
  and both sides are positive, so their prime multiplicities are defined.  Counting factors of \(2\) yields
  \[
    2\nu_2(A)=\nu_2(A^2)
      =\nu_2(2B^2)=1+2\nu_2(B).
  \]
  The left side is even and the right side is odd, a contradiction.  Hence
  \(x^2=2\) has no solutions.
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  We proved that
  \begin{itemize}
    \item division with remainder is always possible, and the pair
      \((q,r)\) is unique;
    \item the Euclidean algorithm computes greatest common divisors and
      yields Bézout's identity;
    \item a prime dividing a product divides one of the factors;
    \item every natural number from \(2\) on factors into primes, uniquely
      up to reordering;
    \item prime multiplicities add on products, so every prime occurs an
      even number of times in a square;
    \item consequently, \(x^2=2\) has no solution in \(\Q\).
  \end{itemize}
\end{frame}

\section{The real numbers}

\subsection{Cauchy sequences of rationals}

\begin{frame}{Why introduce the real numbers?}
  We proved that the equation
  \[
    x^2=2
  \]
  has no solution in \(\Q\).  Nevertheless, rational numbers can approximate
  the desired positive solution with arbitrarily small error.

  We therefore seek a larger ordered field in which
  \begin{itemize}
    \item every rational number remains visible;
    \item limits of rational Cauchy sequences exist;
    \item arithmetic and order are compatible with taking limits.
  \end{itemize}

  \begin{alertblock}{Construction}
    A real number will be an equivalence class of Cauchy sequences of rational
    numbers.  Two sequences represent the same real when their difference
    tends to zero.
  \end{alertblock}
\end{frame}

\begin{frame}{Absolute value}
  \small
  In any linearly ordered field, define
  \[
    \boxed{|x|=\max\{x,-x\}.}
  \]

  We shall use the following standard properties without reproving them.
  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.48\textwidth}
      \begin{block}{Order and symmetry}
        \begin{align*}
          0&\le|x|,\\
          |x|=0&\Longleftrightarrow x=0,\\
          -|x|&\le x\le|x|,\\
          |-x|&=|x|,\\
          |x-y|&=|y-x|.
        \end{align*}
      \end{block}

      For \(\varepsilon\ge0\),
      \[
        |x|\le\varepsilon
        \Longleftrightarrow
        -\varepsilon\le x\le\varepsilon.
      \]
    \end{column}
    \begin{column}{0.48\textwidth}
      \begin{block}{Arithmetic and estimates}
        \begin{align*}
          |xy|&=|x|\,|y|,\\
          |x^{-1}|&=|x|^{-1} &&(x\ne0),\\
          |x+y|&\le|x|+|y|,\\
          \bigl||x|-|y|\bigr|&\le|x-y|.
        \end{align*}
      \end{block}

      For \(\delta\ge0\),
      \[
        \delta<|x|
        \Longleftrightarrow
        x<-\delta\ \text{ or }\ \delta<x.
      \]
    \end{column}
  \end{columns}
\end{frame}

\begin{frame}{Cauchy sequences and pre-reals}
  Let \(x=(x_n)_{n\in\N}\) be a sequence in \(\Q\).

  \begin{block}{Definition}
    The sequence \(x\) is \alert{Cauchy} if
    \[
      \forall\varepsilon>0\ \exists N\ \forall p,q\ge N,
      \qquad |x_p-x_q|\le\varepsilon.
    \]
    Here \(\varepsilon\) is rational.
  \end{block}

  The condition asks the tail of the sequence to have arbitrarily small
  diameter.  It does not mention a limit, since that limit may not lie in
  \(\Q\).

  \begin{block}{Definition}
    A \alert{pre-real} is a rational Cauchy sequence.  We write
    \[
      \R_{\mathrm{pre}}
        =\{x\colon\N\to\Q:x\text{ is Cauchy}\}.
    \]
  \end{block}
\end{frame}

\begin{frame}{Every Cauchy sequence is bounded: the finite part}
  Let \(x\) be Cauchy.  Apply the definition with \(\varepsilon=1\).  There is
  \(A\in\N\) such that
  \[
    p,q\ge A
    \quad\Longrightarrow\quad
    |x_p-x_q|\le1.
  \]

  The initial set
  \[
    \{|x_0|,|x_1|,\ldots,|x_A|\}
  \]
  is finite and nonempty, so it has a maximum \(M\).  In particular,
  \[
    0\le M,
    \qquad
    |x_n|\le M\quad(0\le n\le A).
  \]

  We claim that \(B=M+1\) bounds the whole sequence.
\end{frame}

\begin{frame}{Every Cauchy sequence is bounded: the tail}
  If \(n\le A\), then
  \[
    |x_n|\le M\le M+1.
  \]
  If \(A\le n\), the triangle inequality and the Cauchy estimate give
  \[
    |x_n|
      =|x_A+(x_n-x_A)|
      \le |x_A|+|x_n-x_A|
      \le M+1.
  \]

  Thus
  \[
    \boxed{B=M+1>0,\qquad |x_n|\le B\text{ for every }n.}
  \]

  This boundedness lemma is the key extra ingredient in proving that products
  and reciprocals of suitable Cauchy sequences are again Cauchy.
\end{frame}

\begin{frame}{When do two pre-reals represent the same number?}
  \begin{block}{Definition}
    For \(x,y\in\R_{\mathrm{pre}}\), define
    \[
      x\sim y
      \quad\Longleftrightarrow\quad
      \forall\varepsilon>0\ \exists N\ \forall n\ge N,
      \qquad |x_n-y_n|\le\varepsilon.
    \]
  \end{block}

  In words, \(x\sim y\) means that \(x_n-y_n\) tends to zero.

  Reflexivity follows from \(|x_n-x_n|=0\).  Symmetry follows from
  \[
    |y_n-x_n|=|x_n-y_n|.
  \]

  Transitivity is the only part requiring an estimate.
\end{frame}

\begin{frame}{Transitivity of the equivalence relation}
  Suppose \(x\sim y\) and \(y\sim z\), and fix \(\varepsilon>0\).  Choose
  thresholds \(N,M\) such that
  \[
    |x_n-y_n|\le\frac\varepsilon2\quad(n\ge N),
    \qquad
    |y_n-z_n|\le\frac\varepsilon2\quad(n\ge M).
  \]

  For \(n\ge\max(N,M)\),
  \begin{align*}
    |x_n-z_n|
      &=|(x_n-y_n)+(y_n-z_n)|\\
      &\le |x_n-y_n|+|y_n-z_n|\\
      &\le \frac\varepsilon2+\frac\varepsilon2
       =\varepsilon.
  \end{align*}

  Hence \(x\sim z\), and \(\sim\) is an equivalence relation on the
  pre-reals.
\end{frame}

\subsection{Arithmetic on pre-reals}

\begin{frame}{Constants, negation, and addition}
  Define operations pointwise:
  \[
    0_n=0,
    \qquad
    1_n=1,
    \qquad
    (-x)_n=-x_n,
    \qquad
    (x+y)_n=x_n+y_n.
  \]
  Constant sequences are Cauchy, and negation preserves every Cauchy
  estimate because
  \[
    |(-x_p)-(-x_q)|=|x_q-x_p|=|x_p-x_q|.
  \]

  For addition, choose \(N,M\) so that the variations of \(x\) and \(y\) are
  at most \(\varepsilon/2\).  If \(p,q\ge\max(N,M)\), then
  \begin{align*}
    |(x_p+y_p)-(x_q+y_q)|
      &\le |x_p-x_q|+|y_p-y_q|\\
      &\le\frac\varepsilon2+\frac\varepsilon2=\varepsilon.
  \end{align*}
  Thus \(-x\) and \(x+y\) are pre-reals.
\end{frame}

\begin{frame}{Addition respects equivalence}
  Suppose \(x\sim x'\) and \(y\sim y'\).  Given \(\varepsilon>0\), choose a
  common threshold after which
  \[
    |x_n-x'_n|\le\frac\varepsilon2,
    \qquad
    |y_n-y'_n|\le\frac\varepsilon2.
  \]

  Beyond that threshold,
  \begin{align*}
    |(x_n+y_n)-(x'_n+y'_n)|
      &=|(x_n-x'_n)+(y_n-y'_n)|\\
      &\le |x_n-x'_n|+|y_n-y'_n|\\
      &\le\varepsilon.
  \end{align*}

  Therefore
  \[
    x+y\sim x'+y'.
  \]
  The same argument, using symmetry of absolute value, shows that
  \(x\sim x'\) implies \(-x\sim-x'\).
\end{frame}

\begin{frame}{Products of Cauchy sequences}
  Let \(|x_n|\le A\) and \(|y_n|\le B\), with \(A,B>0\).  Choose a common
  threshold such that
  \[
    |x_p-x_q|\le\frac\varepsilon{2B},
    \qquad
    |y_p-y_q|\le\frac\varepsilon{2A}.
  \]

  Use the identity
  \[
    x_py_p-x_qy_q
      =x_p(y_p-y_q)+y_q(x_p-x_q).
  \]
  Hence
  \begin{align*}
    |x_py_p-x_qy_q|
      &\le |x_p|\,|y_p-y_q|+|y_q|\,|x_p-x_q|\\
      &\le A\frac\varepsilon{2A}+B\frac\varepsilon{2B}
       =\varepsilon.
  \end{align*}

  Thus the pointwise product \((xy)_n=x_ny_n\) is Cauchy.
\end{frame}

\begin{frame}{Multiplication respects equivalence}
  Suppose \(x\sim x'\) and \(y\sim y'\).  Choose bounds
  \[
    |x_n|\le A,
    \qquad
    |y'_n|\le B,
    \qquad A,B>0.
  \]
  The comparison of the two products is organized as
  \[
    x_ny_n-x'_ny'_n
      =x_n(y_n-y'_n)+y'_n(x_n-x'_n).
  \]

  Once
  \[
    |y_n-y'_n|\le\frac\varepsilon{2A},
    \qquad
    |x_n-x'_n|\le\frac\varepsilon{2B},
  \]
  we obtain
  \[
    |x_ny_n-x'_ny'_n|
      \le A\frac\varepsilon{2A}+B\frac\varepsilon{2B}
      =\varepsilon.
  \]
  Therefore \(xy\sim x'y'\).
\end{frame}

\begin{frame}{A nonzero pre-real stays away from zero}
  The condition \(x\not\sim0\) means that for some \(\delta>0\), no tail
  satisfies \(|x_n|\le\delta\).  Thus
  \[
    \exists\delta>0\ \forall N\ \exists M\ge N,
    \qquad \delta<|x_M|.
  \]

  Apply the Cauchy property with \(\delta/2\), obtaining a threshold \(N\).
  Choose \(M\ge N\) with \(|x_M|>\delta\).  For every \(n\ge M\),
  \begin{align*}
    |x_n|
      &=|x_M-(x_M-x_n)|\\
      &\ge |x_M|-|x_M-x_n|\\
      &>\delta-\frac\delta2=\frac\delta2.
  \end{align*}

  Hence there are \(A>0\) and \(M\) such that
  \[
    \boxed{A<|x_n|\quad\text{for every }n\ge M.}
  \]
\end{frame}

\begin{frame}{Reciprocals of nonzero pre-reals are Cauchy}
  Let \(x\not\sim0\).  By the preceding lemma, choose \(A>0\) and \(N\) with
  \(A<|x_n|\) for \(n\ge N\).  Also choose \(M\) so that
  \[
    |x_p-x_q|\le\varepsilon A^2
    \qquad(p,q\ge M).
  \]

  For \(p,q\ge\max(N,M)\), both denominators are nonzero and
  \begin{align*}
    \left|\frac1{x_p}-\frac1{x_q}\right|
      &=\frac{|x_q-x_p|}{|x_p|\,|x_q|}\\
      &\le\frac{|x_q-x_p|}{A^2}\\
      &\le\frac{\varepsilon A^2}{A^2}=\varepsilon.
  \end{align*}

  Thus \((1/x_n)\) is Cauchy whenever the class of \(x\) is nonzero.
\end{frame}

\begin{frame}{The reciprocal is independent of representatives}
  Suppose \(x\sim x'\), with neither sequence equivalent to zero.  Choose
  eventual lower bounds
  \[
    A<|x_n|,
    \qquad
    A'<|x'_n|.
  \]
  Since \(x\sim x'\), eventually
  \[
    |x_n-x'_n|\le\varepsilon AA'.
  \]

  Beyond a common threshold,
  \begin{align*}
    \left|\frac1{x_n}-\frac1{x'_n}\right|
      &=\frac{|x'_n-x_n|}{|x_n|\,|x'_n|}\\
      &\le\frac{|x'_n-x_n|}{AA'}\\
      &\le\varepsilon.
  \end{align*}

  Therefore \((1/x_n)\sim(1/x'_n)\).  Reciprocal is consequently well
  defined on every nonzero equivalence class.
\end{frame}

\subsection{The quotient field}

\begin{frame}{Definition of the real numbers}
  We have proved that \(\sim\) is an equivalence relation on
  \(\R_{\mathrm{pre}}\).

  \begin{block}{Definition}
    The real numbers are the quotient
    \[
      \boxed{\R=\R_{\mathrm{pre}}/{\sim}.}
    \]
    We write \(X=[x]\) when the real number \(X\) is represented by the
    Cauchy sequence \(x\).
  \end{block}

  If \(X=[x]\) and \(Y=[y]\), the operations descend from pointwise
  operations:
  \[
    X+Y=[x+y],
    \qquad
    -X=[-x],
    \qquad
    XY=[xy].
  \]
  If \(X=[x]\ne0\), choose \(N\) with \(x_n\ne0\) for \(n\ge N\).  Its
  reciprocal is represented by any rational sequence whose tail is
  \((1/x_n)_{n\ge N}\); the finitely many earlier values do not affect its
  class.  The estimates already proved show that every formula is independent
  of representatives.
\end{frame}

\begin{frame}{The commutative-ring laws}
  Write \(X=[x]\), \(Y=[y]\), and \(Z=[z]\).  Every ring law holds pointwise
  before passing to the quotient.  For example,
  \[
    ((x+y)+z)_n=(x_n+y_n)+z_n
      =x_n+(y_n+z_n)=(x+(y+z))_n,
  \]
  so
  \[
    (X+Y)+Z=X+(Y+Z).
  \]

  The same representative calculation proves
  \begin{itemize}
    \item associativity and commutativity of addition and multiplication;
    \item the zero and one identity laws;
    \item additive inverses;
    \item distributivity.
  \end{itemize}

  Thus \(\R\) is a commutative ring.
\end{frame}

\begin{frame}{The real numbers are nontrivial}
  The zero and one of \(\R\) are represented by the constant sequences
  \[
    (0,0,0,\ldots),
    \qquad
    (1,1,1,\ldots).
  \]

  If these classes were equal, then for \(\varepsilon=1/2\) there would be a
  threshold \(N\) such that
  \[
    |0-1|\le\frac12
    \qquad(n\ge N).
  \]
  But \(|0-1|=1\), a contradiction.  Therefore
  \[
    \boxed{0\ne1\text{ in }\R.}
  \]

  Hence the quotient contains at least two distinct elements.
\end{frame}

\begin{frame}{Multiplying by a reciprocal}
  Let \(X=[x]\ne0\).  Then \(x\not\sim0\), so there are \(\delta>0\) and
  \(N\) such that
  \[
    \delta<|x_n|
    \qquad(n\ge N).
  \]
  In particular, \(x_n\ne0\) for \(n\ge N\).  On this tail,
  \[
    x_n\frac1{x_n}=1.
  \]

  Hence the product sequence differs from the constant sequence \(1\) only
  in finitely many positions.  Their difference is eventually zero, so they
  are equivalent:
  \[
    \boxed{XX^{-1}=1\qquad(X\ne0).}
  \]

  Therefore the constructed real numbers form a field.
\end{frame}

\begin{frame}{Embedding the rational numbers}
  A rational number determines a constant Cauchy sequence.  Define
  \[
    k\colon\Q\longrightarrow\R,
    \qquad
    k(r)=[(r,r,r,\ldots)].
  \]
  Pointwise arithmetic gives
  \begin{align*}
    k(0)&=0, & k(1)&=1, & k(-r)&=-k(r),\\
    k(r+s)&=k(r)+k(s),
      & k(rs)&=k(r)k(s).
  \end{align*}

  Thus \(k\) is a field homomorphism.  It remains to check that distinct
  rationals do not become equivalent as constant sequences.
\end{frame}

\begin{frame}{The rational embedding is injective}
  Suppose \(k(r)=k(s)\).  Then the two constant sequences are equivalent:
  \[
    \forall\varepsilon>0\ \exists N\ \forall n\ge N,
    \qquad |r-s|\le\varepsilon.
  \]

  If \(r\ne s\), then \(|r-s|>0\).  Choose
  \[
    \varepsilon=\frac{|r-s|}{2}>0.
  \]
  Equivalence would give
  \[
    |r-s|\le\frac{|r-s|}{2},
  \]
  which is impossible.  Hence \(r=s\), so \(k\) is injective.

  We may therefore regard \(\Q\) as a subfield of \(\R\).
\end{frame}

\subsection{Positivity and order}

\begin{frame}{A positive pre-real}
  \begin{block}{Definition}
    A pre-real \(x\) is \alert{positive}, written \(\operatorname{Pos}(x)\), if
    \[
      \exists\delta>0\ \exists N\ \forall n\ge N,
      \qquad \delta\le x_n.
    \]
  \end{block}

  A positive pre-real is not equivalent to zero.  Indeed, if \(x\sim0\),
  fix the positive rational \(\delta/2\).  At a common large index we would have
  \[
    \delta\le x_n
    \qquad\text{and}\qquad
    |x_n|\le\frac\delta2,
  \]
  which is impossible.
\end{frame}

\begin{frame}{Why term-by-term sign conditions do not work}
  \small
  One might try to declare
  \[
    x>0\iff x_n>0\text{ for every }n,
    \qquad
    x\ge0\iff x_n\ge0\text{ for every }n.
  \]
  Neither condition gives a well-defined property of an equivalence class.

  \begin{enumerate}
    \item \emph{Changing finitely many terms changes the answer.}
      The equivalent pre-reals
      \[
        (1,1,1,\ldots)\sim(-1,1,1,\ldots)
      \]
      do not both have all terms positive.  Likewise,
      \((0,0,0,\ldots)\sim(-1,0,0,\ldots)\), but only the first has all
      terms nonnegative.

    \item \emph{Even requiring the sign only eventually is not enough.}
      Set \(u_n=1/(n+1)\).  Then every \(u_n>0\), but \(u\sim0\); the naive
      rule would therefore make the zero class positive.  Also \(-u\sim0\),
      although every term of \(-u\) is negative, so termwise nonnegativity
      would depend on the chosen representative of zero.
  \end{enumerate}

  A fixed eventual gap separates positive classes from zero.  Nonnegative
  will mean: equivalent to zero or positive.
\end{frame}

\begin{frame}{Positivity is independent of representatives}
  Suppose \(x\sim x'\) and \(x_n\ge\delta>0\) for \(n\ge N\).  Choose \(M\)
  such that
  \[
    |x_n-x'_n|\le\frac\delta2
    \qquad(n\ge M).
  \]

  For \(n\ge\max(N,M)\),
  \begin{align*}
    x'_n
      &=x_n-(x_n-x'_n)\\
      &\ge\delta-|x_n-x'_n|\\
      &\ge\delta-\frac\delta2=\frac\delta2.
  \end{align*}

  Hence \(\operatorname{Pos}(x')\).  By symmetry, positivity depends only on
  the equivalence class represented by the sequence.
\end{frame}

\begin{frame}{Positive pre-reals are closed under arithmetic}
  Suppose eventually
  \[
    x_n\ge A>0,
    \qquad
    y_n\ge B>0.
  \]
  Beyond the maximum of the two thresholds,
  \[
    x_n+y_n\ge A+B>0,
  \]
  so \(x+y\) is positive.

  Since all four quantities are nonnegative,
  \[
    x_ny_n\ge AB>0,
  \]
  so \(xy\) is positive as well.

  \begin{block}{Conclusion}
    \[
      \operatorname{Pos}(x),\operatorname{Pos}(y)
      \Longrightarrow
      \operatorname{Pos}(x+y),\operatorname{Pos}(xy).
    \]
  \end{block}
\end{frame}

\begin{frame}{Nonnegative pre-reals}
  \begin{block}{Definition}
    A pre-real \(x\) is \alert{nonnegative} if
    \[
      \operatorname{Nonneg}(x)
      \quad\Longleftrightarrow\quad
      \operatorname{Pos}(x)\ \text{ or }\ x\sim0.
    \]
  \end{block}

  This property is invariant under equivalence because both alternatives are.
  It includes every eventually pointwise nonnegative pre-real.  Indeed, if
  \(x_n\ge0\) for every \(n\ge N\), then:
  \begin{itemize}
    \item if \(x\sim0\), it is nonnegative by definition;
    \item otherwise, \(|x_n|>\delta\) eventually for some \(\delta>0\), and
      the pointwise sign forces \(x_n>\delta\).
  \end{itemize}
  Thus \(x\) is positive in the second case.
\end{frame}

\begin{frame}{Opposite nonnegative signs meet only at zero}
  Suppose both \(x\) and \(-x\) are nonnegative.  If \(x\not\sim0\), then
  also \(-x\not\sim0\), so both must be positive.  There would be
  \(\delta,\delta'>0\) and a common large index \(n\) with
  \[
    x_n\ge\delta,
    \qquad
    -x_n\ge\delta'.
  \]
  Adding yields
  \[
    0\ge\delta+\delta'>0,
  \]
  a contradiction.  Therefore
  \[
    \boxed{\operatorname{Nonneg}(x),\operatorname{Nonneg}(-x)
      \Longrightarrow x\sim0.}
  \]

  This is the antisymmetry mechanism for the real order.
\end{frame}

\begin{frame}{If a pre-real is not nonnegative}
  Assume \(x\) is not nonnegative.  Then \(x\not\sim0\) and \(x\) is not
  positive.

  From \(x\not\sim0\), choose \(\delta>0\) and \(N\) such that
  \[
    |x_n|>\delta\qquad(n\ge N).
  \]
  From the Cauchy property, choose \(M\) such that
  \[
    |x_p-x_q|\le\frac\delta2
    \qquad(p,q\ge M).
  \]

  Since \(x\) is not positive, the proposed gap \(\delta\) fails after every
  threshold.  Hence there is \(R\ge\max(N,M)\) with \(x_R<\delta\).  Together
  with \(|x_R|>\delta\), this forces
  \[
    x_R<-\delta,
    \qquad\text{so}\qquad
    -x_R>\delta.
  \]
\end{frame}

\begin{frame}{The opposite pre-real is positive}
  Continue with the index \(R\) for which \(-x_R>\delta\).  If
  \(n\ge\max(R,N,M)\), the Cauchy estimate gives
  \[
    x_R-x_n\ge-\frac\delta2.
  \]
  Therefore
  \begin{align*}
    -x_n
      &=(x_R-x_n)+(-x_R)\\
      &>-\frac\delta2+\delta
       =\frac\delta2.
  \end{align*}

  Thus \(-x\) is positive, and in particular nonnegative.  We have proved
  the sign alternative
  \[
    \boxed{\operatorname{Nonneg}(x)
      \quad\text{or}\quad
      \operatorname{Nonneg}(-x).}
  \]
  This is the key step toward totality of the order.
\end{frame}

\begin{frame}{Definition of the real order}
  Nonnegativity passes to equivalence classes.  For \(X,Y\in\R\), define
  \[
    \boxed{X\le Y
      \quad\Longleftrightarrow\quad
      Y-X\text{ is nonnegative}.}
  \]

  Reflexivity follows because \(X-X=0\).  For transitivity, if \(Y-X\) and
  \(Z-Y\) are nonnegative, their sum is nonnegative, and
  \[
    (Z-Y)+(Y-X)=Z-X.
  \]

  For antisymmetry, if \(X\le Y\) and \(Y\le X\), then both \(Y-X\) and
  \(-(Y-X)=X-Y\) are nonnegative.  The preceding lemma gives
  \[
    Y-X=0,
  \]
  hence \(X=Y\).
\end{frame}

\begin{frame}{Totality and compatibility with arithmetic}
  Apply the sign alternative to \(Y-X\).  If \(Y-X\) is nonnegative, then
  \(X\le Y\).  Otherwise \(-(Y-X)=X-Y\) is nonnegative, so \(Y\le X\).
  Thus the order is total.

  Translation preserves order because
  \[
    (Y+T)-(X+T)=Y-X.
  \]
  Products of nonnegative classes are nonnegative by the corresponding
  pre-real result.  Products of positive classes are positive.

  \begin{block}{Theorem}
    The constructed real numbers form a linearly ordered field.
  \end{block}
\end{frame}

\begin{frame}{Eventual inequalities are sufficient}
  The eventual non-strict sign condition is not necessary, but it is a useful
  sufficient condition:
  \[
    \boxed{x_n\ge0\text{ eventually}\quad\Longrightarrow\quad [x]\ge0.}
  \]
  Indeed, if \(x\sim0\), then \([x]=0\).  Otherwise, \(|x_n|>\delta\)
  eventually for some \(\delta>0\); combined with \(x_n\ge0\), this gives
  \(x_n>\delta\) eventually, so \(x\) is positive.

  Consequently, if \(X=[x]\) and \(Y=[y]\), then
  \[
    \boxed{x_n\le y_n\text{ eventually}\quad\Longrightarrow\quad X\le Y,}
  \]
  because \(y_n-x_n\ge0\) eventually.

  The converses fail.  For example, \(z_n=-1/(n+1)\) satisfies \(z\sim0\),
  so \([z]=0\ge0\), although no term of \(z\) is nonnegative.  Likewise,
  if \(x_n=0\) and \(y_n=z_n\), then \([x]=[y]\), but \(x_n\le y_n\)
  never holds.
\end{frame}

\begin{frame}{The rational embedding preserves order}
  For \(r,s\in\Q\), the difference \(k(s)-k(r)\) is represented by the
  constant sequence \(s-r\).

  If \(r<s\), this sequence is positive with the fixed gap \(s-r\).  If
  \(r=s\), it represents zero.  Hence
  \[
    r\le s\quad\Longrightarrow\quad k(r)\le k(s).
  \]

  Conversely, if \(s-r<0\), the constant sequence \(s-r\) is neither
  positive nor equivalent to zero, so it cannot be nonnegative.  Therefore
  \[
    \boxed{k(r)\le k(s)\quad\Longleftrightarrow\quad r\le s.}
  \]
  Injectivity then also gives
  \(k(r)<k(s)\Longleftrightarrow r<s\).
\end{frame}

\subsection{Rational density and convergence}

\begin{frame}[label=cauchy-tail-distance]{A Cauchy tail controls the real distance}
  \small
  Let \(X=[x]\in\R\), let \(\varepsilon\in\Q\) be positive, and suppose \(N\)
  works in the Cauchy condition for \(x\) at tolerance \(\varepsilon\):
  \[
    p,q\ge N
    \quad\Longrightarrow\quad
    |x_p-x_q|\le\varepsilon.
  \]
  Fixing \(q=N\), we obtain
  \[
    -\varepsilon\le x_n-x_N\le\varepsilon
    \qquad(n\ge N).
  \]

  Now \(k(x_N)\) is represented by the constant sequence
  \((x_N,x_N,\ldots)\).  Therefore
  \[
    X-k(x_N)=[(x_n-x_N)_{n\in\N}].
  \]
  The eventual inequalities pass to the real order, giving
  \[
    -k(\varepsilon)\le X-k(x_N)\le k(\varepsilon)
  \]
  hence
  \[
\boxed{|X-k(x_N)|\le k(\varepsilon).}
  \]
\end{frame}

\begin{frame}{A positive rational below a positive real}
  Suppose \(X=[x]>0\).  Positivity of its representative gives
  \(\delta>0\) and \(N\) such that
  \[
    x_n\ge\delta
    \qquad(n\ge N).
  \]
  Set \(r=\delta/2\).  Then \(r>0\), and eventually
  \[
    x_n-r\ge\frac\delta2>0.
  \]
  Hence
  \[
    0<k(r)<X.
  \]
  (The positive gap \(\delta/2\) makes \(k(r)<X\) strict.)
\end{frame}

\begin{frame}{The rationals are dense among the reals}
  Let \(X\in\R\), and let \(E>0\) be real.  The previous lemma, applied to
  \(E\), provides a rational \(\varepsilon>0\) with
  \[
    0<k(\varepsilon)<E.
  \]
  The
  \hyperlink{cauchy-tail-distance}{\textcolor{blue}{\underline{Cauchy-tail estimate}}}
  at tolerance \(\varepsilon\) then provides \(r\in\Q\) with
  \[
    |X-k(r)|\le k(\varepsilon)<E.
  \]

  Thus every real number can be approximated by a rational number with
  arbitrarily small real error:
  \[
    \boxed{\forall E>0\ \exists r\in\Q,\qquad |X-k(r)|<E.}
  \]
\end{frame}

\begin{frame}{The Archimedean bound in \(\R\)}
  Let \(X=[x]\).  Boundedness of the pre-real \(x\) gives a positive rational
  \(B\) such that
  \[
    |x_m|\le B\qquad\text{for every }m.
  \]
  \emph{Convention.}  From now on we write \(n\) for the rational number
  \(i(n)\) whenever a natural number appears inside a rational or real
  expression; carrying the embeddings explicitly would only clutter the
  estimates.

  The Archimedean property of \(\Q\) gives \(n\in\N\) with \(B\le n+1\).
  Therefore
  \[
    x_m\le|x_m|\le B\le n+1
  \]
  for every \(m\).  Passing to classes gives
  \[
    \boxed{X\le k(n+1).}
  \]

  Thus the real construction inherits an Archimedean upper bound directly
  from the boundedness of rational Cauchy sequences.
\end{frame}

\begin{frame}{Arbitrarily small reciprocal bounds}
  Let \(E>0\).  Apply the Archimedean bound to \(|E^{-1}|\).  For some
  \(n\in\N\),
  \[
    |E^{-1}|\le k(n+1).
  \]
  Since \(E>0\), its reciprocal is positive, so the left side equals
  \(E^{-1}\).  Both sides are positive; reversing the inequality under
  reciprocals yields
  \[
    \boxed{k\!\left(\frac1{n+1}\right)\le E.}
  \]

  This is the form of the Archimedean property used in both Cauchy estimates
  below: the rational numbers \(1/(n+1)\) eventually fit below every positive
  real number.
\end{frame}

\begin{frame}{Convergence and Cauchy sequences in \(\R\)}
  For a sequence \(F\colon\N\to\R\), define
  \[
    F_n\longrightarrow X
  \]
  to mean
  \[
    \forall E>0\ \exists N\ \forall n\ge N,
    \qquad |F_n-X|\le E.
  \]

  The sequence is \alert{Cauchy} if
  \[
    \forall E>0\ \exists N\ \forall p,q\ge N,
    \qquad |F_p-F_q|\le E.
  \]

  The real numbers are \alert{complete} if every real Cauchy sequence
  converges to a real number.  The construction was designed so that a
  rational Cauchy sequence converges to the class it represents.
\end{frame}

\begin{frame}{Rational tolerances are enough}
  Let \(F\colon\N\to\R\) and \(X\in\R\).  Suppose that for every positive
  rational \(\varepsilon\) we can prove
  \[
    \exists N\ \forall n\ge N,
    \qquad |F_n-X|\le k(\varepsilon).
  \]
  Then \(F_n\to X\) for arbitrary positive real tolerances.

  Indeed, given \(E>0\), choose a positive rational \(\delta\) with
  \[
    0<k(\delta)<E.
  \]
  The hypothesis at \(\delta\) supplies \(N\) such that, for \(n\ge N\),
  \[
    |F_n-X|\le k(\delta)<E,
  \]
  and in particular \(|F_n-X|\le E\).

  We will repeatedly use this criterion below.
\end{frame}

\begin{frame}{A representative sequence converges: the idea}
  \small
  Let \(x=(x_n)_{n\in\N}\) represent \(X=[x]\).  Applying the rational
  embedding term by term gives the real sequence
  \[
    k(x_0),\ k(x_1),\ k(x_2),\ldots
  \]
  We claim that this sequence converges to \(X\).  Each \(k(x_n)\) is
  represented by the constant row obtained by \emph{freezing} \(x_n\),
  whereas \(X\) is represented by the moving bottom row:
  \[
    \begin{array}{c|ccccc}
      \text{real number}
        & \multicolumn{5}{c}{\text{a rational representative}}\\
      \hline
      k(x_0) & x_0 & x_0 & x_0 & x_0 & \cdots\\
      k(x_1) & x_1 & x_1 & x_1 & x_1 & \cdots\\
      k(x_2) & x_2 & x_2 & x_2 & x_2 & \cdots\\
      \vdots & \vdots&\vdots&\vdots&\vdots&\\
      k(x_n) & x_n & x_n & x_n & x_n & \cdots\\
      \hline
      X=[x]             & x_0 & x_1 & x_2 & x_3 & \cdots
    \end{array}
  \]

  Thus \(k(x_n)\to X\) means that, as we freeze terms farther along the
  sequence, the constant rows get closer to the moving bottom row.  The
  Cauchy property says exactly that late terms stay close to one another,
  and that is all the claim requires.
\end{frame}

\begin{frame}{A representative sequence converges to its class}
  \small
  Let \(x\in\R_{\mathrm{pre}}\) and set \(X=[x]\).  We claim
  \[
    k(x_n)\longrightarrow X.
  \]

  Fix a positive rational \(\varepsilon\).  The Cauchy property
  supplies \(N\) such that for \(n,m\ge N\),
  \[
    |x_n-x_m|\le\varepsilon.
  \]
  In the table, this says that beyond column \(N\) every frozen row with
  \(n\ge N\) agrees with the moving row to within \(\varepsilon\); the
  finitely many earlier columns are irrelevant, because the quotient order
  only compares tails.  As explained
  \hyperlink{cauchy-tail-distance}{\textcolor{blue}{\underline{above}}},
  this gives
  \[
    |k(x_n)-X|\le k(\varepsilon)
    \qquad(n\ge N).
  \]

  We have verified the rational-tolerance hypothesis for every
  \(\varepsilon>0\) in \(\Q\).  The preceding reduction therefore gives
  \[
    \boxed{k(x_n)\longrightarrow X.}
  \]
\end{frame}

\subsection{Completeness}

\begin{frame}{Completeness: statement and strategy}
  \begin{block}{Theorem}
    Every Cauchy sequence of real numbers converges to a real number.
  \end{block}

  \medskip
  Let \(F\colon\N\to\R\) be a Cauchy sequence.  The proof has three steps.
  \begin{enumerate}
    \item \alert{Approximate.}  Rational density provides, for each \(n\),
      a rational \(x_n\) with
      \(\bigl|F_n-k(x_n)\bigr|<k\bigl(\tfrac1{n+1}\bigr)\).
    \item \alert{Transfer the Cauchy property.}  The rational sequence
      \(x=(x_n)\) is Cauchy because \(F\) is; it is therefore a pre-real
      and defines a real number \(X=[x]\).
    \item \alert{Pass to the limit.}  We already know that
      \(k(x_n)\to X\), and the approximation errors tend to zero; hence
      \(F_n\to X\).
  \end{enumerate}
\end{frame}

\begin{frame}{Approximating a real sequence term by term}
  Let \(F\colon\N\to\R\).  Rational density gives, for every \(n\), a rational
  \(x_n\) satisfying
  \[
    \boxed{|F_n-k(x_n)|
      <k\!\left(\frac1{n+1}\right).}
  \]

  Choosing one such rational for each \(n\) defines a sequence
  \[
    x=\operatorname{approx}(F)\colon\N\longrightarrow\Q.
  \]

  Any tolerances shrinking to zero would serve; the choice \(1/(n+1)\)
  guarantees that the error at stage \(n\) tends to zero, so \(F_n\) and
  \(k(x_n)\) will have the same limiting behaviour.

  The goal is to prove that \(F_n\) converges to the real number
  represented by \(x\).  To speak of that number at all, we must first
  prove that \(x\) is a rational Cauchy sequence whenever \(F\) is a real
  Cauchy sequence.
\end{frame}

\begin{frame}{The rational approximation is Cauchy}
  \footnotesize
  Assume \(F\) is Cauchy, and fix a positive rational \(\varepsilon\).  Choose
  \(N\) such that
  \[
    |F_p-F_q|\le\frac{k(\varepsilon)}3
    \qquad(p,q\ge N),
  \]
  and use the Archimedean property to choose \(M\) such that
  \[
    k\!\left(\frac1{M+1}\right)
      \le\frac{k(\varepsilon)}3.
  \]

  For \(p\ge M\), we have \(M+1\le p+1\); taking reciprocals reverses this
  comparison, so the approximation bound gives
  \[
    |F_p-k(x_p)|
      <k\!\left(\frac1{p+1}\right)
      \le k\!\left(\frac1{M+1}\right)
      \le\frac{k(\varepsilon)}3.
  \]

  If \(p,q\ge\max(M,N)\), splitting \(k(x_p)-k(x_q)\) into three
  differences gives
  \begin{align*}
    |k(x_p)-k(x_q)|
      &\le |k(x_p)-F_p|+|F_p-F_q|+|F_q-k(x_q)|\\
      &\le\frac{k(\varepsilon)}3
        +\frac{k(\varepsilon)}3
        +\frac{k(\varepsilon)}3
       =k(\varepsilon).
  \end{align*}
  Since \(k\) reflects order, \(|x_p-x_q|\le\varepsilon\).  Thus \(x\) is a
  rational Cauchy sequence.
\end{frame}

\begin{frame}{Every real Cauchy sequence converges}
  \footnotesize
  The rational approximation \(x\) is a pre-real, so let
  \[
    X=[x]\in\R.
  \]
  We already know that \(k(x_n)\to X\).

  Fix a positive real \(\varepsilon\).  Choose \(N\) so that
  \[
    |k(x_n)-X|\le\frac{\varepsilon}{2}
    \qquad(n\ge N),
  \]
  and use the Archimedean property to choose \(M\) so that
  \[
    k\!\left(\frac1{M+1}\right)\le\frac{\varepsilon}{2}.
  \]
  For \(n\ge M\), reciprocals reverse the comparison \(M+1\le n+1\), so the
  approximation bound gives
  \[
    |F_n-k(x_n)|
      <k\!\left(\frac1{n+1}\right)
      \le k\!\left(\frac1{M+1}\right)
      \le\frac{\varepsilon}{2}.
  \]
  For \(n\ge\max(N,M)\), combining the two halves,
  \begin{align*}
    |F_n-X|
      &\le |F_n-k(x_n)|+|k(x_n)-X|\\
      &\le\frac{\varepsilon}{2}+\frac{\varepsilon}{2}
       =\varepsilon.
  \end{align*}
  Hence \(F_n\to X\).
\end{frame}

\begin{frame}{The real numbers are complete}
  \begin{block}{Theorem}
    Every Cauchy sequence of real numbers converges to a real number.
  \end{block}

  \medskip
  Looking back at the three steps:
  \begin{itemize}
    \item rational density produced the approximations \(x_n\), with errors
      controlled by \(1/(n+1)\);
    \item the Cauchy hypothesis on \(F\) was used exactly once: to prove
      that the rational approximation \(x\) is itself Cauchy;
    \item the convergence \(k(x_n)\to X\) came directly from the
      construction of \(\R\), and the vanishing errors transferred it to
      \(F\).
  \end{itemize}

  \medskip
  The construction has repaired the defect that motivated it: Cauchy
  sequences that had no limit in \(\Q\) acquire one in \(\R\).
\end{frame}

\subsection{\texorpdfstring{A solution of \(x^2=2\)}{A solution of x squared equals 2}}

\begin{frame}{Newton's rational approximations}
  \small
  Define a rational sequence \(a=(a_n)\) recursively by
  \[
    a_0=1,
    \qquad
    a_{n+1}=\frac12\left(a_n+\frac2{a_n}\right).
  \]
  The idea: if \(a_n^2\) falls short of \(2\), then \((2/a_n)^2\) overshoots
  it, and vice versa, so the average of \(a_n\) and \(2/a_n\) should
  approximate the desired number better than either.

  The sequence is well defined and positive.  Indeed, \(a_0>0\), and if
  \(a_n>0\), then both \(a_n\) and \(2/a_n\) are positive, so their average
  is positive.  The first few terms are
  \[
    1,\qquad
    \frac32,\qquad
    \frac{17}{12},\qquad
    \frac{577}{408},\quad\ldots
  \]

  \begin{alertblock}{The plan}
    Show that the errors \(a_n^2-2\) tend to zero, deduce that \((a_n)\) is
    Cauchy, and prove that the real number it represents squares to \(2\).
  \end{alertblock}
\end{frame}

\begin{frame}{The exact error formula}
  \small
  Define the rational error sequence \(e=(e_n)\) by
  \[
    e_n=a_n^2-2.
  \]
  Direct calculation gives
  \begin{align*}
    e_{n+1}
      &=\frac14\left(a_n+\frac2{a_n}\right)^2-2\\
      &=\frac{a_n^2+4+4/a_n^2-8}{4}\\
      &=\boxed{\frac{(a_n^2-2)^2}{4a_n^2}}
       =\frac{e_n^2}{4a_n^2}.
  \end{align*}

  The right side is a square divided by a positive number, so
  \(e_{n+1}\ge0\) for every \(n\): from index \(1\) on, the approximations
  overshoot,
  \[
    a_n^2\ge2\qquad(n\ge1).
  \]
  Also \(a_1=3/2\), hence \(e_1=1/4\).
\end{frame}

\begin{frame}{The errors tend to zero}
  \small
  For \(n\ge1\), we have \(a_n^2\ge2\), so
  \[
    0\le e_{n+1}
      =\frac{e_n^2}{4a_n^2}
      \le\frac{e_n^2}{8}.
  \]
  Starting from \(e_1=1/4\), induction gives \(e_n\le1/4\).  Consequently,
  \[
    e_{n+1}\le\frac{e_n}{32},
  \]
  and another induction yields
  \[
    \boxed{0\le e_n\le\frac1{4\cdot32^{n-1}}\qquad(n\ge1).}
  \]

  These upper bounds tend to zero.  Given a rational \(\varepsilon>0\), the
  Archimedean property provides \(m\) with \(m+1\ge1/(4\varepsilon)\), and
  \(32^m\ge m+1\) by induction, so
  \[
    e_n\le\frac1{4\cdot32^{m}}\le\frac1{4(m+1)}\le\varepsilon
    \qquad(n\ge m+1).
  \]
\end{frame}

\begin{frame}{The Newton sequence is Cauchy}
  \footnotesize
  Since \(a_n>0\) and \(a_n^2\ge2\), we have \(a_n\ge1\) for \(n\ge1\):
  a number below \(1\) would have square below \(1\).  Hence, from index
  \(1\) on, the sequence decreases by a controlled amount:
  \[
    0\le a_n-a_{n+1}
      =\frac{a_n^2-2}{2a_n}
      =\frac{e_n}{2a_n}
      \le\frac{e_n}{2}.
  \]

  If \(q>p\ge1\), the differences telescope, and iterating
  \(e_{j+1}\le e_j/32\) bounds them by a geometric series:
  \begin{align*}
    |a_p-a_q|
      &=\sum_{j=p}^{q-1}(a_j-a_{j+1})
       \le\frac12\sum_{j=p}^{q-1}e_j\\
      &\le\frac{e_p}{2}\sum_{i=0}^{q-p-1}\frac1{32^i}
       <\frac{16}{31}\,e_p<e_p.
  \end{align*}

  Given a rational \(\varepsilon>0\), choose \(N\ge1\) with
  \(e_N\le\varepsilon\).  The errors decrease, so for \(q>p\ge N\),
  \[
    |a_p-a_q|<e_p\le e_N\le\varepsilon.
  \]
  Thus \((a_n)\) is Cauchy.
\end{frame}

\begin{frame}{A real number whose square is \(2\)}
  \small
  The rational sequence \(a=(a_n)\) is a pre-real, so define
  \[
    \alpha=[a]\in\R.
  \]
  Since \(a_n\ge1\), the sequence has a fixed positive gap and \(\alpha>0\).

  Pointwise multiplication gives \(\alpha^2=[(a_n^2)]\), so
  \[
    \alpha^2-2=[e].
  \]
  Given a rational \(\varepsilon>0\), the error estimate supplies
  \(N\ge1\) such that for \(n\ge N\),
  \[
    |e_n|=e_n\le\varepsilon,
  \]
  the errors being nonnegative from index \(1\) on.  Thus \(e\sim0\), and
  therefore
  \[
    \boxed{\alpha^2=2.}
  \]

  No rational has this property, so \(\alpha\) is a genuinely new number in
  the Cauchy completion of \(\Q\).
\end{frame}

\begin{frame}{The equation \(x^2+1=0\) in \(\R\)}
  Suppose that \(x\in\R\).  Since \(\R\) is an ordered field, every square
  is nonnegative:
  \[
    x^2\ge0.
  \]
  Adding \(1\) and using \(1>0\) gives
  \[
    x^2+1\ge1>0.
  \]
  In particular, \(x^2+1\) cannot equal \(0\).  Therefore
  \[
    \boxed{x^2+1=0\text{ has no solution in }\R.}
  \]

  Completeness fills the gaps detected by Cauchy sequences, but it does not
  guarantee solutions to polynomial equations.  We therefore need to enlarge
  \(\R\) once more.
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  We proved that
  \begin{itemize}
    \item pre-reals are rational Cauchy sequences, modulo differences tending
      to zero;
    \item boundedness controls products, and a nonzero class stays uniformly
      away from zero, which controls reciprocals;
    \item the quotient \(\R\) is a linearly ordered field containing an
      order-preserving copy of \(\Q\);
    \item rational points are dense in \(\R\);
    \item every real Cauchy sequence converges;
    \item the Newton Cauchy sequence defines a positive
      \(\alpha\in\R\) satisfying \(\alpha^2=2\);
    \item nevertheless, \(x^2+1=0\) has no solution in \(\R\).
  \end{itemize}
\end{frame}

\section{The complex numbers}

\subsection{Definition and field structure}

\begin{frame}{Why introduce the complex numbers?}
  We have just proved that the equation
  \[
    x^2+1=0
  \]
  has no solution in \(\R\).  Equivalently, no real number has square
  \(-1\), even though \(\R\) is complete.

  We enlarge \(\R\) by adjoining a symbol \(i\) satisfying
  \[
    \boxed{i^2=-1.}
  \]
  The resulting number system will
  \begin{itemize}
    \item be a field containing \(\R\);
    \item carry an absolute value and be complete;
    \item contain a root of every nonconstant polynomial with complex
      coefficients.
  \end{itemize}
\end{frame}

\begin{frame}{Definition of the complex numbers}
  \begin{block}{Definition}
    The set of complex numbers is
    \[
      \boxed{\C=\R^2.}
    \]
    We write the pair \((a,b)\) as \(a+bi\), where
    \[
      i=(0,1).
    \]
  \end{block}

  Equality is componentwise:
  \[
    a+bi=c+di
    \quad\Longleftrightarrow\quad
    a=c\ \text{ and }\ b=d.
  \]
  Thus every complex number has a unique \emph{real part} and
  \emph{imaginary part},
  \[
    \operatorname{Re}(a+bi)=a,
    \qquad
    \operatorname{Im}(a+bi)=b.
  \]
  We identify \(a\in\R\) with \(a+0i\in\C\).
\end{frame}

\begin{frame}{Addition and multiplication}
  For \(z=a+bi\) and \(w=c+di\), define
  \begin{align*}
    z+w&=(a+c)+(b+d)i,\\
    zw&=(ac-bd)+(ad+bc)i.
  \end{align*}
  The multiplication formula is exactly what formal expansion and
  \(i^2=-1\) predict:
  \[
    (a+bi)(c+di)
      =ac+(ad+bc)i+bd\,i^2.
  \]

  The distinguished elements and additive inverse are
  \[
    0=0+0i,
    \qquad
    1=1+0i,
    \qquad
    -(a+bi)=(-a)+(-b)i.
  \]
  In particular,
  \[
    i^2=(0,1)(0,1)=(-1,0)=-1.
  \]
  Restricted to the pairs \((a,0)\), both operations reduce to real
  addition and multiplication, so the identification of \(\R\) inside
  \(\C\) preserves arithmetic.
\end{frame}

\begin{frame}{The commutative-ring laws}
  Addition is componentwise, so its associativity and commutativity follow
  immediately from the corresponding laws in \(\R\).

  Substituting the multiplication formula shows that
  \[
    zw=wz,
    \qquad
    (zw)u=z(wu),
    \qquad
    z(w+u)=zw+zu,
    \qquad
    1z=z.
  \]
  Each real and imaginary component reduces to an identity in \(\R\).  For
  example, writing \(u=e+fi\), both real components in the distributive law
  are
  \[
    a(c+e)-b(d+f)=(ac-bd)+(ae-bf),
  \]
  and the imaginary components agree in the same way.

  Also \(0\ne1\), because the pairs \((0,0)\) and \((1,0)\) are distinct.
  Hence these operations make \(\C\) a nontrivial commutative ring.
\end{frame}

\begin{frame}{Conjugation and multiplicative inverses}
  Define the \alert{complex conjugate} of \(z=a+bi\) by
  \[
    \overline z=a-bi.
  \]
  Direct calculation gives
  \[
    z\overline z=(a+bi)(a-bi)=a^2+b^2.
  \]
  If \(z\ne0\), then \(a\) and \(b\) are not both zero.  Since squares are
  nonnegative in \(\R\), this implies
  \[
    a^2+b^2>0.
  \]
  We may therefore define
  \[
    \boxed{z^{-1}=\frac{\overline z}{z\overline z}
      =\frac{a}{a^2+b^2}-\frac{b}{a^2+b^2}i.}
  \]
  Then \(zz^{-1}=1\).  Every nonzero element has an inverse, so
  \begin{alertblock}{Conclusion}
    The complex numbers form a field containing \(\R\).
  \end{alertblock}
\end{frame}

\begin{frame}{The complex field cannot be ordered}
  Suppose that a linear order made \(\C\) into an ordered field.  In any
  ordered field every square is nonnegative.  Indeed, if \(x\ge0\), then
  \(x^2\ge0\); if \(x<0\), then \(-x>0\) and \(x^2=(-x)^2\ge0\).

  Applied to \(i\), this would give
  \[
    -1=i^2\ge0.
  \]
  Adding \(1\) to both sides yields \(0\ge1\).  But \(1=1^2\ge0\) and
  \(1\ne0\), so \(0<1\) in any ordered field --- a contradiction.  Therefore
  \[
    \boxed{\text{no linear order makes \(\C\) an ordered field}.}
  \]

  This does not prevent us from measuring the \emph{size} of a complex
  number; for that we use the real order.
\end{frame}

\subsection{Absolute value and completeness}

\begin{frame}{The complex absolute value}
  A standard consequence of the completeness of \(\R\) is that every
  nonnegative real number \(r\) has a unique nonnegative square root
  \(\sqrt r\); the Newton iteration used for \(x^2=2\) adapts to any
  \(r\ge0\).

  \begin{block}{Definition}
    For \(z=a+bi\in\C\), define
    \[
      \boxed{|z|=\sqrt{a^2+b^2} \in\R.}
    \]
  \end{block}
  Equivalently,
  \[
    |z|^2=z\overline z,
    \qquad
    |z|\ge0.
  \]
  It follows at once that
  \[
    |z|=0\Longleftrightarrow z=0,
    \qquad
    |\overline z|=|z|,
    \qquad
    |\operatorname{Re}z|\le|z|,
    \qquad
    |\operatorname{Im}z|\le|z|.
  \]
\end{frame}

\begin{frame}{Multiplicativity of the absolute value}
  Conjugation respects multiplication:
  \[
    \overline{zw}=\overline z\,\overline w.
  \]
  Hence
  \begin{align*}
    |zw|^2
      &=(zw)\overline{(zw)}\\
      &=z\overline z\,w\overline w\\
      &=|z|^2|w|^2
       =\bigl(|z|\,|w|\bigr)^2.
  \end{align*}
  Both \(|zw|\) and \(|z|\,|w|\) are nonnegative, so uniqueness of the
  nonnegative square root gives
  \[
    \boxed{|zw|=|z|\,|w|.}
  \]
  In particular, for \(z\ne0\),
  \[
    |z^{-1}|=|z|^{-1}.
  \]
\end{frame}

\begin{frame}{The triangle inequality}
  \small
  For any complex number \(u\), the coordinate formula gives
  \[
    \operatorname{Re}(u)\le|u|.
  \]
  Therefore
  \begin{align*}
    |z+w|^2
      &=(z+w)(\overline z+\overline w)\\
      &=|z|^2+2\operatorname{Re}(z\overline w)+|w|^2\\
      &\le|z|^2+2|z\overline w|+|w|^2\\
      &=|z|^2+2|z|\,|w|+|w|^2
       =\bigl(|z|+|w|\bigr)^2,
  \end{align*}
  by multiplicativity: \(|z\overline w|=|z|\,|\overline w|=|z|\,|w|\).
  Both sides before squaring are nonnegative, so
  \[
    \boxed{|z+w|\le|z|+|w|.}
  \]
  Thus
  \[
    d(z,w)=|z-w|
  \]
  is a metric on \(\C\).
\end{frame}

\begin{frame}{Convergent and Cauchy sequences}
  Let \(z=(z_n)_{n\in\N}\) be a sequence in \(\C\).

  \begin{block}{Convergence}
    We say that \(z\) \alert{converges} to \(w\in\C\) if
    \[
      \forall\varepsilon>0\ \exists N\ \forall n\ge N,
      \qquad |z_n-w|\le\varepsilon.
    \]
  \end{block}

  \begin{block}{Cauchy condition}
    We say that \(z\) is \alert{Cauchy} if
    \[
      \forall\varepsilon>0\ \exists N\ \forall p,q\ge N,
      \qquad |z_p-z_q|\le\varepsilon.
    \]
  \end{block}
  Here \(\varepsilon\) is always a positive real number.  The field \(\C\)
  is called \alert{complete} if every complex Cauchy sequence converges in
  \(\C\).
\end{frame}

\begin{frame}{Every convergent sequence is Cauchy}
  Suppose \(z_n\to z\), and let \(\varepsilon>0\).  There is an
  \(N\) such that
  \[
    |z_n-z|\le\frac\varepsilon2
    \qquad(n\ge N).
  \]
  If \(p,q\ge N\), the triangle inequality gives
  \begin{align*}
    |z_p-z_q|
      &=|(z_p-z)+(z-z_q)|\\
      &\le|z_p-z|+|z_q-z|\\
      &\le\frac\varepsilon2+\frac\varepsilon2
       =\varepsilon.
  \end{align*}
  Hence \((z_n)\) is Cauchy.

  The converse is precisely the completeness question.  We now reduce it to
  the completeness of the two real coordinates.
\end{frame}

\begin{frame}{A complex Cauchy sequence has Cauchy coordinates}
  Let
  \[
    z_n=a_n+b_ni
  \]
  be Cauchy in \(\C\).  Since
  \[
    |a_p-a_q|
      \le |z_p-z_q|,
    \qquad
    |b_p-b_q|
      \le |z_p-z_q|,
  \]
  the real sequences \((a_n)\) and \((b_n)\) are Cauchy.

  The completeness of \(\R\) supplies real numbers \(a,b\) such that
  \[
    a_n\longrightarrow a,
    \qquad
    b_n\longrightarrow b.
  \]
  The only possible complex limit is therefore
  \[
    z=a+bi
  \]
  (uniqueness of limits is left as an exercise).
  It remains to check convergence in the complex absolute value.
\end{frame}

\begin{frame}{The complex numbers are complete}
  Fix \(\varepsilon>0\).  The two coordinate convergences give an \(N\) such
  that, for \(n\ge N\),
  \[
    |a_n-a|\le\frac\varepsilon2,
    \qquad
    |b_n-b|\le\frac\varepsilon2.
  \]
  For such \(n\), the triangle inequality and \(|ri|=|r|\) give
  \begin{align*}
    |z_n-(a+bi)|
      &=|(a_n-a)+(b_n-b)i|\\
      &\le|a_n-a|+|(b_n-b)i|\\
      &=|a_n-a|+|b_n-b|\\
      &\le\varepsilon.
  \end{align*}
  Thus \(z_n\to a+bi\).

  \begin{alertblock}{Completeness of \(\C\)}
    Every complex Cauchy sequence converges: the completeness of \(\R\),
    applied to the two coordinate sequences, proves the completeness of
    \(\C\).
  \end{alertblock}
\end{frame}

\subsection{Polynomial equations}

\begin{frame}{The equations \(x^2+1=0\) and \(x^4+1=0\)}
  By construction,
  \[
    i^2+1=0,
  \]
  so \(x^2+1=0\) has the solutions \(i\) and \(-i\) in \(\C\).

  Set
  \[
    \alpha=\frac{1+i}{\sqrt2}.
  \]
  Since \((\sqrt2)^2=2\),
  \[
    \alpha^2
      =\frac{(1+i)^2}{2}
      =\frac{1+2i+i^2}{2}
      =i,
    \qquad
    \alpha^4=i^2=-1.
  \]
  Hence \(\alpha^4+1=0\).  In fact the four displayed choices
  \[
    \boxed{\frac{1+i}{\sqrt2},\quad
      \frac{-1-i}{\sqrt2},\quad
      \frac{1-i}{\sqrt2},\quad
      \frac{-1+i}{\sqrt2}}
  \]
  all satisfy \(x^4+1=0\).
\end{frame}

\begin{frame}{The Fundamental Theorem of Algebra}
  Let
  \[
    P(z)=a_0+a_1z+\cdots+a_nz^n,
    \qquad
    a_n\ne0,
    \quad n\ge1,
  \]
  be a nonconstant polynomial with complex coefficients.  The Fundamental
  Theorem of Algebra states that there exists \(\zeta\in\C\) such that
  \[
    \boxed{P(\zeta)=0.}
  \]

  We state this theorem without proof.  Despite its name it is not a purely
  algebraic fact: every proof uses the completeness of \(\R\) in an
  essential way.

  \begin{block}{Consequence}
    Dividing out one root at a time, every polynomial of degree \(n\)
    factors completely:
    \[
      P(z)=a_n(z-\zeta_1)\cdots(z-\zeta_n).
    \]
  \end{block}
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  We proved that
  \begin{itemize}
    \item \(\C=\R^2\), with multiplication determined by \(i^2=-1\), is a
      field containing \(\R\);
    \item no linear order can make \(\C\) an ordered field;
    \item \(|a+bi|=\sqrt{a^2+b^2}\) is multiplicative and satisfies the
      triangle inequality;
    \item convergence and the Cauchy condition are measured by this absolute
      value;
    \item the completeness of \(\R\) in each coordinate proves that \(\C\)
      is complete;
    \item \(x^2+1=0\) and \(x^4+1=0\) have explicit complex solutions.
  \end{itemize}

  We also stated, without proof, the Fundamental Theorem of Algebra: every
  nonconstant complex polynomial has a root in \(\C\).
\end{frame}

\section{\texorpdfstring{The \(p\)-adic numbers}{The p-adic numbers}}

\subsection{Infinite expansions to the left}

\begin{frame}{How an infinite decimal expansion works}
  A real number is reached through its truncations.  For example,
  \[
    538,\quad
    538.2,\quad
    538.21,\quad
    538.219,\quad
    538.2194,\quad\ldots
    \quad\longrightarrow\quad
    538.21947\ldots
  \]
  Each step glues one more digit onto the \emph{right}, changing the value
  by at most \(9\cdot10^{-k}\) with \(k\) growing.  The truncations form a
  Cauchy sequence precisely because high \emph{negative} powers of \(10\)
  are small: digits far to the right barely matter.

  \begin{alertblock}{The question behind this section}
    That is a choice, not a necessity.  What if we instead decide that
    digits far to the \emph{left} barely matter?
  \end{alertblock}
\end{frame}

\begin{frame}{Growing to the left instead}
  Run the same process in the opposite direction:
  \[
    3.706,\quad
    53.706,\quad
    153.706,\quad
    4153.706,\quad
    84153.706,\quad\ldots
    \quad\longrightarrow\quad
    \ldots84153.706
  \]
  Each step glues one more digit onto the \emph{left}.  Writing \(a_j\)
  for the digit multiplying \(10^j\), the step from the truncation
  \(x_k\) to a later truncation \(x_\ell\) adds only digits in positions
  \(k\le j<\ell\):
  \[
    x_\ell-x_k
      =\sum_{j=k}^{\ell-1}a_j\,10^j
      =10^k\cdot(\text{an integer}).
  \]

  \begin{alertblock}{The guiding idea}
    In a distance where high \emph{positive} powers of \(10\) are small,
    these truncations also form a Cauchy sequence, and the left-infinite
    string denotes its limit.
  \end{alertblock}
\end{frame}

\begin{frame}{Real versus \(n\)-adic expansions}
  The same idea works in any base \(n\ge2\), with digits
  \(0,1,\ldots,n-1\).

  \begin{columns}[T,onlytextwidth]
    \begin{column}{0.47\textwidth}
      \begin{block}{Real expansions}
        A real base-\(n\) expansion has only finitely many digits
        \emph{left} of the point, but may continue forever to the right:
        \[
          \pm a_m\cdots a_1a_0
          .a_{-1}a_{-2}a_{-3}\cdots
        \]
      \end{block}
    \end{column}
    \begin{column}{0.47\textwidth}
      \begin{alertblock}{\(n\)-adic expansions}
        An \(n\)-adic expansion may continue forever to the
        \emph{left}, but has only finitely many digits to the right:
        \[
          \cdots a_3a_2a_1a_0
          .a_{-1}\cdots a_{-s}
        \]
      \end{alertblock}
    \end{column}
  \end{columns}

  Each system allows infinitely many digits on exactly one side of the
  point.  In base \(10\), compare
  \[
    \pm538.21947\ldots
    \qquad\text{with}\qquad
    \ldots84153.706.
  \]
  Note the missing \(\pm\) on the right: an \(n\)-adic expansion carries
  no sign.  We will see shortly why none is needed.
\end{frame}

\begin{frame}{Arithmetic still runs from right to left}
  In a left-infinite decimal expansion, addition is performed by the usual
  digit algorithm:
  \[
    \begin{array}{r}
      \ldots84153.706\\
      {}+\ \ldots70926.153\\ \hline
      \ldots55079.859
    \end{array}
  \]

  Each requested final block of digits depends on only finitely many
  positions to its right.  Equivalently, for every \(k\), ordinary arithmetic
  determines the answer modulo \(10^k\).

  Multiplication works in the same way: compute to any chosen finite
  precision, and increasing the precision does not change the digits already
  determined near the point.

  \begin{block}{Digit strings behave better than real expansions}
    Two left-infinite strings are equal exactly when \emph{all} their
    digits agree; there is no analogue of \(0.999\ldots=1.000\ldots\).
    Moreover, each digit of a sum or product depends on finitely many
    digits of the inputs.  For real expansions this fails: in
    \(0.333\ldots+0.666\ldots\), no finite number of digits decides
    whether the integer part is \(0\) or \(1\).
  \end{block}
\end{frame}

\begin{frame}{Negative numbers acquire infinite left tails}
  \small
  The finite decimal strings
  \[
    9,\qquad99,\qquad999,\qquad9999,\quad\ldots
  \]
  approach \(-1\) in the \(10\)-adic sense.  Indeed,
  \[
    (10^k-1)-(-1)=10^k,
  \]
  and a high power of \(10\) will have very small \(10\)-adic size.  Thus
  \[
    \boxed{-1=\ldots999999.}
  \]
  The digit algorithm agrees: adding \(1\) to \(\ldots999999\) launches a
  carry that travels forever to the left, so
  \(\ldots999999+1=\ldots000000=0\).

  Similarly,
  \[
    83,\qquad983,\qquad9983,\qquad99983,\quad\ldots
  \]
  approach \(-17\), because \(10^k-17\) differs from \(-17\) by \(10^k\).
  Hence
  \[
    \boxed{-17=\ldots999983.}
  \]

  This explains the missing sign: a left-infinite string needs no
  \(\pm\), because negatives are already ordinary digit strings --- the
  infinite tail of \(9\)s does the work of the minus sign.
\end{frame}

\begin{frame}{A rational left-infinite expansion}
  \small
  Solving \(7x=3\) modulo successive powers of \(10\) gives
  \[
    9,\quad29,\quad429,\quad1429,\quad71429,\quad571429,\quad8571429,\ldots
  \]
  For instance, \(7\cdot1429=10003\) and \(7\cdot8571429=60000003\): the
  products agree with \(3\) in the final four and seven digits.

  These compatible solutions converge \(10\)-adically to
  \[
    \boxed{\frac37=\ldots8571428571429.}
  \]
  The digit algorithm confirms this: multiplying \(\ldots8571429\) by \(7\)
  right to left starts with \(7\cdot9=63\), and every later step produces a
  multiple of \(10\), so the product is \(\ldots0000003=3\).

  Dividing by \(10\) merely shifts the point:
  \[
    \frac3{70}=\ldots857142857142.9.
  \]

  Rational numbers therefore appear as eventually periodic left-infinite
  expansions, just as they appear as eventually periodic real expansions to
  the right.
\end{frame}

\begin{frame}{Which digits control distance?}
  In the ordinary metric, agreement far to the \emph{right} of the point
  means closeness.  For example,
  \[
    |8.21947\ldots-8.21929\ldots|
      =0.00018\ldots<10^{-3}.
  \]

  In the \(10\)-adic metric, closeness means agreement in the fractional
  digits and in many final digits \emph{left} of the point:
  \[
    \begin{aligned}
      x&=\ldots84153.7,\\
      y&=\ldots26153.7.
    \end{aligned}
    \qquad
    x-y=\ldots58000.0.
  \]
  Their difference is divisible by \(10^3\) and by no higher power of
  \(10\), so their \(10\)-adic distance will be
  \[
    \|x-y\|_{10}=10^{-3}.
  \]

  This is the bridge from digit strings to the formal definition: count the
  powers of the base dividing a difference.
\end{frame}

\subsection{\texorpdfstring{The \(p\)-adic size}{The p-adic size}}

\begin{frame}{From base \(10\) to a prime base}
  The opening slides used base \(10\) because its digits are familiar.
  For the formal theory we now fix a \alert{prime} \(p\) and work in base
  \(p\).

  \begin{block}{Why a prime base?}
    Everything below can be generalized to an arbitrary base \(n\ge2\),
    but the prime case is simpler: divisibility is governed by a single
    prime, and the resulting size is \emph{multiplicative}.
  \end{block}

  From the arithmetic section: \(\nu_p(m)\) counts the factors \(p\) in
  the unique prime factorization of \(m\), and
  \[
    \nu_p(m_1m_2)=\nu_p(m_1)+\nu_p(m_2).
  \]
  We set \(\nu_p(-m)=\nu_p(m)\), so that \(\nu_p\) is defined on all
  nonzero integers.
\end{frame}

\begin{frame}{The \(p\)-adic order}
  \begin{block}{Definition}
    For \(x\in\Q^\times\), write \(x=a/b\) with nonzero integers \(a,b\)
    and define the \alert{\(p\)-adic order}
    \[
      \boxed{\nu_p(x)=\nu_p(a)-\nu_p(b),
      \qquad
      \nu_p(0)=+\infty.}
    \]
  \end{block}

  This is well defined: if \(a/b=c/d\), then \(ad=bc\), and additivity
  gives
  \[
    \nu_p(a)+\nu_p(d)=\nu_p(b)+\nu_p(c),
  \]
  so the two differences agree.

  Additivity persists on \(\Q^\times\):
  \[
    \nu_p(xy)=\nu_p(x)+\nu_p(y).
  \]
\end{frame}

\begin{frame}{Order measures common final digits}
  For a nonzero integer \(a\) and \(k\in\N\), directly from unique
  factorization,
  \[
    \boxed{p^k\mid a\quad\Longleftrightarrow\quad \nu_p(a)\ge k.}
  \]

  In particular,
  \[
    \nu_p(p^k)=k,
  \]
  and \(\nu_p(a)\) is exactly the largest exponent \(k\) for which
  \(p^k\mid a\).

  If two integers have the same final \(k\) base-\(p\) digits, their
  difference is divisible by \(p^k\), and hence has \(p\)-adic order at
  least \(k\).
\end{frame}

\begin{frame}{The \(p\)-adic size}
  \begin{block}{Definition}
    For \(x\in\Q\), define
    \[
      \boxed{
      \|x\|_p=
      \begin{cases}
        0,&x=0,\\
        p^{-\nu_p(x)},&x\ne0.
      \end{cases}}
    \]
  \end{block}

  The digit interpretation becomes
  \[
    p^k\mid a
    \quad\Longleftrightarrow\quad
    \|a\|_p\le p^{-k}
    \qquad(a\in\Z).
  \]
  For example, with \(p=5\),
  \[
    \|58000\|_5=5^{-3},
    \qquad
    \|5^k\|_5=5^{-k},
    \qquad
    \|58001\|_5=1.
  \]

  This confirms the intuition: adding base-\(p\) digits farther to the
  left changes a truncation by a \(p\)-adically smaller amount.
\end{frame}

\begin{frame}{The order laws}
  If \(p^m\) divides two integers, it divides their sum.  Writing
  \(x=a/c\) and \(y=b/c\) over a common denominator, this yields
  \[
    \boxed{\nu_p(x+y)\ge\min\{\nu_p(x),\nu_p(y)\}.}
  \]
  Indeed \(\nu_p(a+b)\ge\min\{\nu_p(a),\nu_p(b)\}\); subtract
  \(\nu_p(c)\) from both sides.

  Multiplication is even better: additivity is an \emph{equality},
  \[
    \boxed{\nu_p(xy)=\nu_p(x)+\nu_p(y).}
  \]
\end{frame}

\begin{frame}{Properties of the \(p\)-adic size}
  Translating the order laws gives
  \[
    \|x\|_p\ge0,
    \qquad
    \|x\|_p=0\Longleftrightarrow x=0,
    \qquad
    \|-x\|_p=\|x\|_p,
  \]
  together with the ultrametric inequality
  \[
    \boxed{\|x+y\|_p
      \le\max\{\|x\|_p,\|y\|_p\}}
  \]
  and multiplicativity
  \[
    \boxed{\|xy\|_p=\|x\|_p\,\|y\|_p.}
  \]

  The usual triangle and reverse-triangle estimates follow:
  \[
    \|x+y\|_p\le\|x\|_p+\|y\|_p,
    \qquad
    \bigl|\|x\|_p-\|y\|_p\bigr|\le\|x-y\|_p.
  \]
  In particular, \(\|x^{-1}\|_p=\|x\|_p^{-1}\) for \(x\ne0\).
\end{frame}

\begin{frame}{The dominant term and the metric}
  The ultrametric inequality has a useful strengthening:
  \[
    \|x\|_p\ne\|y\|_p
    \quad\Longrightarrow\quad
    \boxed{\|x+y\|_p=\max\{\|x\|_p,\|y\|_p\}.}
  \]
  Indeed, suppose \(\nu_p(x)<\nu_p(y)\).  The sum law gives
  \(\nu_p(x+y)\ge\nu_p(x)\); applied to \(x=(x+y)-y\), it gives
  \(\nu_p(x)\ge\min\{\nu_p(x+y),\nu_p(y)\}\), and since
  \(\nu_p(y)>\nu_p(x)\) this forces \(\nu_p(x+y)=\nu_p(x)\).

  Define
  \[
    d_p(x,y)=\|x-y\|_p.
  \]
  Then \(d_p\) is an ultrametric:
  \[
    d_p(x,z)\le\max\{d_p(x,y),d_p(y,z)\}.
  \]

  We now complete \(\Q\) with respect to this distance.
\end{frame}

\subsection{Cauchy sequences and the quotient}

\begin{frame}{\(p\)-adic Cauchy sequences}
  Let \(x=(x_k)_{k\in\N}\) be a sequence of rational numbers.

  \begin{block}{Definition}
    The sequence \(x\) is \alert{\(p\)-adically Cauchy} if
    \[
      \forall\varepsilon>0\ \exists N\ \forall k,\ell\ge N,
      \qquad \|x_k-x_\ell\|_p\le\varepsilon.
    \]
  \end{block}

  Write
  \[
    \mathcal C_p
      =\{x\colon\N\to\Q:x\text{ is \(p\)-adically Cauchy}\}.
  \]

  The truncations of a left-infinite base-\(p\) string lie in
  \(\mathcal C_p\): if \(\ell>k\), their difference is divisible by
  \(p^k\), so
  \[
    \|x_\ell-x_k\|_p\le p^{-k}\longrightarrow0.
  \]
\end{frame}

\begin{frame}{Cauchy sequences are \(p\)-adically bounded}
  Let \(x\in\mathcal C_p\).  Choose \(N\ge1\) such that
  \[
    \|x_k-x_\ell\|_p\le1
    \qquad(k,\ell\ge N).
  \]
  For \(k\ge N\), the ultrametric inequality gives
  \[
    \|x_k\|_p
      \le\max\{\|x_k-x_N\|_p,\|x_N\|_p\}
      \le\max\{1,\|x_N\|_p\}.
  \]

  Taking the maximum of this bound and the finitely many initial sizes gives
  \(B>0\) such that
  \[
    \boxed{\|x_k\|_p\le B\qquad\text{for every }k.}
  \]

  This is the same boundedness argument used for real Cauchy sequences, with
  the ultrametric maximum replacing a sum.
\end{frame}

\begin{frame}{Equivalent Cauchy sequences}
  \begin{block}{Definition}
    For \(x,y\in\mathcal C_p\), define
    \[
      x\sim_p y
      \quad\Longleftrightarrow\quad
      \|x_k-y_k\|_p\longrightarrow0.
    \]
  \end{block}

  Reflexivity and symmetry are immediate.  If \(x\sim_p y\) and
  \(y\sim_p z\), then
  \[
    \|x_k-z_k\|_p
      \le\max\{\|x_k-y_k\|_p,\|y_k-z_k\|_p\}
      \longrightarrow0.
  \]
  Thus \(\sim_p\) is an equivalence relation.

  Two digit strings represent the same class precisely when their
  truncations eventually agree to every fixed finite precision.
\end{frame}

\begin{frame}{Definition of the \(p\)-adic numbers}
  \begin{block}{Definition}
    The set of \alert{\(p\)-adic numbers} is
    \[
      \boxed{\Q_p=\mathcal C_p/{\sim_p}.}
    \]
    We write \(X=[x]\) when \(X\) is represented by the \(p\)-adic Cauchy
    sequence \(x\).
  \end{block}

  Define constants and arithmetic pointwise:
  \[
    (-x)_k=-x_k,
    \qquad
    (x+y)_k=x_k+y_k,
    \qquad
    (xy)_k=x_ky_k.
  \]
  Therefore
  \[
    -[x]=[-x],
    \qquad
    [x]+[y]=[x+y],
    \qquad
    [x][y]=[xy].
  \]
\end{frame}

\begin{frame}{Why arithmetic descends to the quotient}
  Negation and addition preserve Cauchy sequences directly.  For products,
  boundedness and multiplicativity give
  \begin{align*}
    \|x_ky_k-x_\ell y_\ell\|_p
      \le\max\bigl\{
        &\|x_k\|_p\,\|y_k-y_\ell\|_p,\\
        &\|y_\ell\|_p\,\|x_k-x_\ell\|_p
        \bigr\}.
  \end{align*}
  The right side tends to zero as \(k,\ell\to\infty\).

  The same estimate proves that equivalent representatives give equivalent
  sums and products.  All commutative-ring laws hold pointwise before
  passing to classes.

  \begin{alertblock}{Conclusion}
    For every prime \(p\), \(\Q_p\) is a commutative ring.
  \end{alertblock}
\end{frame}

\begin{frame}{The rational numbers inside \(\Q_p\)}
  A rational number determines a constant Cauchy sequence.  Define
  \[
    \iota_p\colon\Q\longrightarrow\Q_p,
    \qquad
    \iota_p(r)=[(r,r,r,\ldots)].
  \]
  Pointwise arithmetic makes \(\iota_p\) a ring homomorphism.

  If \(\iota_p(r)=\iota_p(s)\), then the constant sequence
  \(\|r-s\|_p\) tends to zero.  If \(r\ne s\), choosing
  \[
    \varepsilon=\frac{\|r-s\|_p}{2}>0
  \]
  makes this impossible.  Therefore \(r=s\), and \(\iota_p\) is injective.

  We may consequently regard \(\Q\) as a subring of \(\Q_p\).  In
  particular, \(\Q_p\) is nontrivial.
\end{frame}

\subsection{\texorpdfstring{\(\Q_p\) is a field}{Qp is a field}}

\begin{frame}{Reciprocals in \(\Q_p\)}
  \small
  Let \(X=[x]\ne0\).  Since \(x\not\sim_p0\), there is \(\delta>0\) with
  \(\|x_k\|_p>\delta\) for infinitely many \(k\); choose \(N\) such that
  \(\|x_k-x_\ell\|_p\le\delta\) for \(k,\ell\ge N\).  The dominant-term
  lemma then freezes the size along the tail: there is \(c>\delta\) with
  \[
    \|x_k\|_p=c\qquad(k\ge N).
  \]
  In particular \(x_k\ne0\) there, and the tail \(y_k=x_k^{-1}\) is Cauchy,
  because multiplicativity gives
  \[
    \|y_k-y_\ell\|_p
      =\frac{\|x_k-x_\ell\|_p}{\|x_k\|_p\,\|x_\ell\|_p}
      =\frac{\|x_k-x_\ell\|_p}{c^2}.
  \]
  The finitely many earlier values are irrelevant; since \(x_ky_k=1\) on
  the tail, the class \(Y=[y]\) satisfies \(XY=1\).

  \begin{alertblock}{Conclusion}
    Every nonzero \(p\)-adic number is invertible:
    \(\Q_p\) is a field containing \(\Q\).
  \end{alertblock}
\end{frame}

\subsection{\texorpdfstring{The extended size and completeness}{The extended size and completeness}}

\begin{frame}{Extending the \(p\)-adic size to \(\Q_p\)}
  If \(x\in\mathcal C_p\), the reverse-triangle estimate gives
  \[
    \bigl|\|x_k\|_p-\|x_\ell\|_p\bigr|
      \le\|x_k-x_\ell\|_p.
  \]
  Thus \((\|x_k\|_p)\) is a Cauchy sequence of real numbers and has a real
  limit.

  \begin{block}{Definition}
    For \(X=[x]\in\Q_p\), define
    \[
      \boxed{\|X\|_p=\lim_{k\to\infty}\|x_k\|_p.}
    \]
  \end{block}

  If \(x\sim_p y\), the same estimate shows that the two real limits agree.
  Hence the definition is independent of the representative.
\end{frame}

\begin{frame}{Size and distance on \(\Q_p\)}
  Passing the rational identities and inequalities to limits gives
  \[
    \|X\|_p\ge0,
    \qquad
    \|X\|_p=0\Longleftrightarrow X=0,
    \qquad
    \|-X\|_p=\|X\|_p,
  \]
  \[
    \|XY\|_p=\|X\|_p\,\|Y\|_p,
    \qquad
    \|X+Y\|_p\le\max\{\|X\|_p,\|Y\|_p\}.
  \]
  In particular, multiplicativity survives the limit.

  Thus
  \[
    d_p(X,Y)=\|X-Y\|_p
  \]
  is an ultrametric on \(\Q_p\).  The rational embedding preserves distance:
  \[
    \|\iota_p(r)-\iota_p(s)\|_p=\|r-s\|_p.
  \]
\end{frame}

\begin{frame}{The rational numbers are dense in \(\Q_p\)}
  Let \(X=[x]\in\Q_p\) and fix \(\varepsilon>0\).  Since \(x\) is Cauchy,
  choose \(N\) such that
  \[
    \|x_k-x_N\|_p\le\frac{\varepsilon}{2}
    \qquad(k\ge N).
  \]
  The difference \(X-\iota_p(x_N)\) is represented by
  \((x_k-x_N)_k\), so
  \[
    \|X-\iota_p(x_N)\|_p
      =\lim_{k\to\infty}\|x_k-x_N\|_p
      \le\frac{\varepsilon}{2}<\varepsilon.
  \]

  Therefore
  \[
    \boxed{\forall X\in\Q_p\ \forall\varepsilon>0\
      \exists q\in\Q,\qquad
      \|X-\iota_p(q)\|_p<\varepsilon.}
  \]
\end{frame}

\begin{frame}{Completeness: rational approximations}
  Let \((X_k)_{k\in\N}\) be a Cauchy sequence in \(\Q_p\), Cauchy and
  convergent sequences in \(\Q_p\) being defined as in \(\C\), with
  \(\|\cdot\|_p\) in place of the absolute value.  By rational density,
  choose \(q_k\in\Q\) such that
  \[
    \|X_k-\iota_p(q_k)\|_p<p^{-k}.
  \]

  The rational sequence \(q=(q_k)\) is \(p\)-adically Cauchy, because
  \begin{align*}
    \|q_k-q_\ell\|_p
      \le\max\bigl\{
        &\|\iota_p(q_k)-X_k\|_p,\,
        \|X_k-X_\ell\|_p,\\
        &\|X_\ell-\iota_p(q_\ell)\|_p
        \bigr\},
  \end{align*}
  and all three terms tend to zero as \(k,\ell\to\infty\).

  Hence \(q\in\mathcal C_p\), and it defines
  \[
    X=[q]\in\Q_p.
  \]
\end{frame}

\begin{frame}{Every Cauchy sequence converges}
  Because \(q=(q_k)\) is Cauchy,
  \[
    \iota_p(q_k)\longrightarrow[q]=X.
  \]
  Indeed, for every \(\varepsilon>0\), all sufficiently large \(k,\ell\)
  satisfy
  \[
    \|q_\ell-q_k\|_p\le\frac{\varepsilon}{2},
  \]
  and therefore
  \[
    \|X-\iota_p(q_k)\|_p
      =\lim_{\ell\to\infty}\|q_\ell-q_k\|_p
      \le\frac{\varepsilon}{2}<\varepsilon.
  \]

  Finally,
  \[
    \|X_k-X\|_p
      \le\max\bigl\{
        \|X_k-\iota_p(q_k)\|_p,\,
        \|\iota_p(q_k)-X\|_p
        \bigr\}\longrightarrow0.
  \]

  \begin{alertblock}{Completeness of \(\Q_p\)}
    Every \(p\)-adic Cauchy sequence in \(\Q_p\) converges in \(\Q_p\).
  \end{alertblock}
\end{frame}

\subsection{\texorpdfstring{The \(p\)-adic integers}{The p-adic integers}}

\begin{frame}{The closed unit ball}
  The field \(\Q_p\) contains elements of arbitrarily large and small
  \(p\)-adic size.  We single out those of size at most \(1\).

  \begin{block}{Definition}
    The \alert{\(p\)-adic integers} form the closed unit ball
    \[
      \boxed{\Z_p
        =\{X\in\Q_p:\|X\|_p\le1\}
        =\overline B(0,1).}
    \]
  \end{block}

  For a rational number \(x\), this condition has the familiar description
  \[
    x\in\Q\cap\Z_p
    \quad\Longleftrightarrow\quad
    \nu_p(x)\ge0.
  \]
  Informally, these are the \(p\)-adic numbers whose base-\(p\) expansion
  has no digits to the right of the point --- a picture made precise at
  the end of this section.
\end{frame}

\begin{frame}{Why the unit ball is a ring}
  Certainly \(0,1\in\Z_p\).  If \(X,Y\in\Z_p\), the ultrametric and
  multiplicative laws give
  \[
    \|X-Y\|_p
      \le\max\{\|X\|_p,\|Y\|_p\}\le1,
    \qquad
    \|XY\|_p=\|X\|_p\|Y\|_p\le1.
  \]
  Thus \(X-Y\) and \(XY\) again lie in \(\Z_p\).

  \begin{alertblock}{Conclusion}
    The unit ball \(\Z_p\) is a subring of the field \(\Q_p\).
  \end{alertblock}

  This is a genuinely \(p\)-adic phenomenon.  The real unit ball is
  \[
    \{x\in\R:|x|\le1\}=[-1,1],
  \]
  which is not a ring: it contains \(1\), but not \(1+1=2\).  The ordinary
  triangle inequality only bounds a sum by \(2\); the ultrametric inequality
  keeps it inside the same unit ball.
\end{frame}

\begin{frame}{The ordinary integers inside \(\Z_p\)}
  If \(n\in\Z\) is nonzero, then \(\nu_p(n)\) counts the copies of \(p\)
  in its prime factorization.  In particular,
  \[
    \nu_p(n)\ge0
    \qquad\Longrightarrow\qquad
    \|n\|_p=p^{-\nu_p(n)}\le1.
  \]
  The same is immediate for \(n=0\).  Hence the embedding
  \(\Q\hookrightarrow\Q_p\) restricts to an inclusion
  \[
    \boxed{\Z\subseteq\Z_p.}
  \]

  There are two useful possibilities:
  \[
    \|n\|_p=
    \begin{cases}
      1,&p\nmid n,\\
      p^{-k},&p^k\mid n\text{ but }p^{k+1}\nmid n.
    \end{cases}
  \]
  Thus divisibility by higher powers of \(p\) makes an ordinary integer
  progressively smaller inside \(\Z_p\).
\end{frame}

\begin{frame}{The rational numbers inside \(\Z_p\)}
  \small
  \begin{block}{Definition}
    Let \(\Z_{(p)}\) be the set of fractions whose denominator is not
    divisible by \(p\):
    \[
      \Z_{(p)}
        =\left\{\frac ab\in\Q:a,b\in\Z,\ p\nmid b\right\}.
    \]
  \end{block}
  A denominator not divisible by \(p\) has \(\nu_p(b)=0\).  For a nonzero
  fraction of this form,
  \[
    \left\|\frac ab\right\|_p
      =p^{-(\nu_p(a)-\nu_p(b))}
      =p^{-\nu_p(a)}\le1,
  \]
  and so
  \[
    \boxed{\Z\subseteq\Z_{(p)}\subseteq\Z_p.}
  \]

  In fact these are exactly the rational points of the unit ball:
  \[
    \boxed{\Q\cap\Z_p=\Z_{(p)}.}
  \]
  Indeed, after a fraction is reduced, a \(p\) in its denominator would
  give it negative \(p\)-adic order.
\end{frame}

\begin{frame}{The \(p\)-adic integers are complete}
  The size map \(X\mapsto\|X\|_p\) is continuous: the reverse-triangle
  estimate, valid in \(\Q_p\) exactly as in \(\Q\), gives
  \[
    \bigl|\|X\|_p-\|Y\|_p\bigr|\le\|X-Y\|_p.
  \]
  Consequently,
  \[
    \Z_p=\{X\in\Q_p:\|X\|_p\le1\}
  \]
  is a closed subspace of \(\Q_p\).

  We already know that \(\Q_p\) is complete, and every closed subspace of a
  complete metric space is complete.  Therefore
  \[
    \boxed{\Z_p\text{ is complete.}}
  \]

  Concretely, a Cauchy sequence of \(p\)-adic integers has a limit in
  \(\Q_p\); closedness says that this limit still has size at most \(1\),
  and hence still belongs to \(\Z_p\).
\end{frame}

\begin{frame}{The ordinary integers are dense in \(\Z_p\)}
  \small
  Let \(X\in\Z_p\) and \(N\in\N\).  Rational density in \(\Q_p\) gives
  \(q\in\Q\) with \(\|X-q\|_p\le p^{-N}\), and the ultrametric inequality
  keeps the approximation inside the unit ball:
  \[
    \|q\|_p\le\max\{\|X\|_p,\|q-X\|_p\}\le1.
  \]
  Hence \(q\in\Q\cap\Z_p=\Z_{(p)}\): write \(q=a/b\) with \(p\nmid b\).

  Since \(b\) and \(p^N\) have no common prime factor, Bézout's identity
  provides \(u,v\in\Z\) with \(ub+vp^N=1\).  The integer \(c=ua\) then
  satisfies
  \[
    q-c=\frac{a(1-ub)}{b}=\frac{av}{b}\,p^N,
    \qquad\text{so}\qquad
    \|q-c\|_p\le p^{-N},
  \]
  and one more ultrametric step gives \(\|X-c\|_p\le p^{-N}\).

  \begin{alertblock}{A non-Archimedean surprise}
    The ordinary integers are dense in \(\Z_p\).  Compare with \(\R\),
    where \(\Z\) is discrete and closed --- as far from dense as a set
    can be.
  \end{alertblock}
\end{frame}

\begin{frame}{The canonical remainder at each level}
  \small
  Fix \(X\in\Z_p\) and \(N\ge1\).  Among all integer approximations of
  \(X\) there is a preferred one: exactly one
  \[
    r_N\in\{0,1,\ldots,p^N-1\}
    \qquad\text{satisfies}\qquad
    \|X-r_N\|_p\le p^{-N}.
  \]

  \emph{Existence.}  Density gives \(c\in\Z\) with \(\|X-c\|_p\le p^{-N}\);
  let \(r_N\) be the remainder of \(c\) upon division by \(p^N\).  The
  discarded multiple of \(p^N\) has size at most \(p^{-N}\), so the
  ultrametric inequality keeps \(r_N\) within \(p^{-N}\) of \(X\).

  \emph{Uniqueness.}  Two candidates differ by an ordinary integer of size
  at most \(p^{-N}\), that is, by a multiple of \(p^N\); both lying in
  \(\{0,\ldots,p^N-1\}\), they are equal.

  Uniqueness also forces \alert{compatibility}: the remainder of
  \(r_{N+1}\) modulo \(p^N\) approximates \(X\) within \(p^{-N}\) as well,
  whence
  \[
    \boxed{r_{N+1}\equiv r_N\pmod{p^N}.}
  \]
\end{frame}

\begin{frame}{From remainders to digits}
  Compatibility means \(r_{N+1}=r_N+a_Np^N\) for a unique digit
  \(a_N\in\{0,1,\ldots,p-1\}\).  The canonical remainders therefore unfold
  digit by digit,
  \[
    r_N=a_0+a_1p+a_2p^2+\cdots+a_{N-1}p^{N-1},
  \]
  and \(\|X-r_N\|_p\le p^{-N}\longrightarrow0\) says exactly that
  \[
    \boxed{X=a_0+a_1p+a_2p^2+a_3p^3+\cdots}
  \]
  as a limit of truncations.

  \begin{alertblock}{The opening picture, now justified}
    Every \(p\)-adic integer is a left-infinite base-\(p\) digit string
    \(\ldots a_3a_2a_1a_0\), and distinct strings represent distinct
    \(p\)-adic integers, by the uniqueness of the canonical remainders.
  \end{alertblock}
\end{frame}

\begin{frame}{A purely arithmetic description of \(\Z_p\)}
  \small
  Collect the canonical remainders of a \(p\)-adic integer into a single
  sequence.  This yields a bijection, which we state without proof:
  \[
    \boxed{\Z_p\;\longleftrightarrow\;
      \bigl\{(r_1,r_2,r_3,\ldots):
        0\le r_n<p^n
        \text{ and }
        r_{n+1}\equiv r_n\pmod{p^n}\bigr\},}
  \]
  under which sums and products are computed \alert{level by level}: add
  or multiply the \(n\)-th remainders, then reduce modulo \(p^n\).

  \medskip
  Why the two sides carry the same information:
  \begin{itemize}
    \item two \(p\)-adic integers with the same remainders at every level
      lie within \(p^{-n}\) of each other for all \(n\), hence coincide;
    \item any compatible sequence satisfies \(r_m\equiv r_n\pmod{p^n}\)
      for \(m\ge n\), so it is \(p\)-adically Cauchy, and its limit is a
      \(p\)-adic integer whose canonical remainders are the given ones;
    \item level-by-level arithmetic is the digit algorithm of the opening
      slides: any finite precision of a sum or product needs only
      finitely many digits of the inputs.
  \end{itemize}

  \medskip
  The right-hand side never mentions distances or Cauchy sequences: a
  \(p\)-adic integer \emph{is} a coherent choice of remainder modulo
  every power of \(p\).
\end{frame}

\subsection{\texorpdfstring{Beyond \(\Q_p\): towards \(\C_p\)}{Beyond Qp: towards Cp}}

\begin{frame}{A polynomial without roots in \(\Q_p\)}
  While constructing reciprocals we saw that for a nonzero \(X=[x]\) the
  sizes \(\|x_k\|_p\) are eventually constant.  The limit defining
  \(\|X\|_p\) is therefore one of these rational sizes:
  \[
    \boxed{\|X\|_p=p^{-m}\ \text{for some }m\in\Z\qquad(X\ne0).}
  \]
  The completion created no new sizes.

  Now suppose \(Y\in\Q_p\) satisfied \(Y^2=p\).  Multiplicativity would
  give
  \[
    \|Y\|_p^2=\|p\|_p=p^{-1},
    \qquad\text{so}\qquad
    \|Y\|_p=p^{-1/2},
  \]
  which is not an integer power of \(p\) --- impossible.

  \begin{alertblock}{\(\Q_p\) is not algebraically closed}
    The polynomial \(x^2-p\) has no root in \(\Q_p\).
  \end{alertblock}

  In \(\R\), the rootless polynomial \(x^2+1\) was detected by the
  \emph{order}: squares are never negative.  In \(\Q_p\), the rootless
  polynomial \(x^2-p\) is detected by the \emph{size}.
\end{frame}

\begin{frame}{Adjoining a square root is not enough}
  \small
  Nothing forbids enlarging \(\Q_p\) exactly as we enlarged \(\R\):
  \[
    \Q_p(\sqrt p)=\{A+B\sqrt p:A,B\in\Q_p\},
  \]
  with multiplication dictated by \((\sqrt p)^2=p\), just as \(i^2=-1\)
  dictated multiplication in \(\C\).  The conjugate trick
  \((A+B\sqrt p)(A-B\sqrt p)=A^2-pB^2\) produces reciprocals --- the
  denominator is nonzero precisely because \(p\) has no square root ---
  and the size extends, with \(\|\sqrt p\|_p=p^{-1/2}\).  Sizes in
  \(\Q_p(\sqrt p)\) are now the powers of \(p^{1/2}\).

  In the real case this single step was miraculous: adjoining one root of
  one polynomial produced \(\C\), where \emph{every} nonconstant
  polynomial has a root.

  \begin{alertblock}{The miracle does not repeat}
    The size argument simply moves one level up: a root of \(x^4-p\)
    would have size \(p^{-1/4}\), which is not a power of \(p^{1/2}\).
    So \(x^4-p\) has no root in \(\Q_p(\sqrt p)\).
  \end{alertblock}

  Adjoining a fourth root of \(p\) repairs this and breaks again at
  \(x^8-p\): no finite number of adjunctions can finish the job.
\end{frame}

\begin{frame}{A gigantic extension}
  Repairing every failure at once requires a leap.  One can construct a
  field \(\overline{\Q_p}\supseteq\Q_p\), gathered from the roots of
  \emph{all} polynomials, such that
  \begin{itemize}
    \item every nonconstant polynomial with coefficients in
      \(\overline{\Q_p}\) has a root in \(\overline{\Q_p}\) --- the field
      is \alert{algebraically closed};
    \item the \(p\)-adic size extends to \(\overline{\Q_p}\), its nonzero
      values now being \emph{all} powers \(p^r\) with \(r\) rational.
  \end{itemize}
  We state its existence without proof.

  \begin{alertblock}{The price of the leap}
    Unlike the two-dimensional step from \(\R\) to \(\C\), this extension
    is infinite-dimensional over \(\Q_p\), and it is \alert{not
    complete}: it contains Cauchy sequences with no limit.
  \end{alertblock}

  Algebra has been repaired, but analysis is broken again.
\end{frame}

\begin{frame}{Completing once more: the field \(\C_p\)}
  We know exactly what to do with an incomplete field carrying a size:
  run the Cauchy construction a third time, now on \(\overline{\Q_p}\).

  \begin{block}{Definition}
    The field of \alert{\(p\)-adic complex numbers} is the completion
    \[
      \boxed{\C_p
        =\bigl(\text{Cauchy sequences in }\overline{\Q_p}\bigr)
        \big/\bigl(\text{differences tending to }0\bigr).}
    \]
  \end{block}

  One might fear an endless alternation --- complete, close
  algebraically, complete again, \(\ldots\)  A remarkable theorem, which
  we state without proof, says that the process stops here.

  \begin{alertblock}{Theorem (without proof)}
    \(\C_p\) is complete \emph{and} algebraically closed.
  \end{alertblock}

  Thus \(\C_p\) stands to \(\Q_p\) as \(\C\) stands to \(\R\): a
  complete, algebraically closed field.
\end{frame}

\subsection{What we have built}

\begin{frame}{Final picture}
  \small
  For a fixed prime \(p\), we proved that
  \begin{itemize}
    \item left-infinite digit strings are understood through compatible
      finite truncations;
    \item \(\nu_p\) counts factors of \(p\), and \(\|x\|_p=p^{-\nu_p(x)}\)
      measures \(p\)-adic size;
    \item this size is ultrametric and multiplicative;
    \item \(\Q_p\) is the quotient of rational \(p\)-adic Cauchy sequences
      by differences tending to zero;
    \item pointwise arithmetic makes \(\Q_p\) a complete field containing
      a dense copy of \(\Q\);
    \item its closed unit ball \(\Z_p\) is a complete subring containing
      \(\Z_{(p)}\), and \(\Z\) is dense in \(\Z_p\);
    \item a \(p\)-adic integer amounts to a compatible choice of
      remainders modulo every power of \(p\) --- equivalently, a
      left-infinite base-\(p\) digit string;
    \item \(x^2-p\) has no root in \(\Q_p\): completeness does not make
      the field algebraically closed.
  \end{itemize}

  \medskip
  We also glimpsed, without proof, the remedy: adjoining the roots of all
  polynomials and completing once more produces \(\C_p\), a field both
  complete and algebraically closed --- the \(p\)-adic counterpart of
  \(\C\).
\end{frame}

\end{document}
